---
title: "pepVet Compared to Other Tools"
description: >
  How pepVet compares to other in-silico digest evaluation tools including
  MS-Digest, ExPASy PeptideMass, Protein Cleaver, ProteaseGuru, and
  PeptideRanger.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{pepVet Compared to Other Tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(pepVet)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r data, include = FALSE}
base_dir <- system.file("extdata", "comparison-data", package = "pepVet")
overlap <- read.csv(file.path(base_dir, "sectionA-overlap.csv"))
caps <- read.csv(file.path(base_dir, "sectionB-capabilities.csv"))
scores <- read.csv(file.path(base_dir, "sectionC-scores-all.csv"))
presets <- read.csv(file.path(base_dir, "sectionC2-presets.csv"))
pr_data <- read.csv(file.path(base_dir, "sectionD-peptideranger.csv"))
pc_data <- read.csv(file.path(base_dir, "sectionE-protein-cleaver.csv"))

enzyme_labels <- c(
  "trypsin" = "Trypsin",
  "lysc" = "Lys-C",
  "chymotrypsin-high" = "Chymotrypsin",
  "glutamyl endopeptidase" = "Glu-C",
  "asp-n endopeptidase" = "Asp-N"
)
enzyme_fill <- c(
  "trypsin" = "#3B7A9E",
  "lysc" = "#6BA292",
  "chymotrypsin-high" = "#D4A76A",
  "glutamyl endopeptidase" = "#C46A6A",
  "asp-n endopeptidase" = "#8B7EB5"
)
verdict_fill <- c(
  "Good" = "#3A8C5F",
  "Moderate" = "#D4A76A",
  "Poor" = "#C46A6A"
)
```

You are choosing an enzyme for a bottom-up proteomics experiment. Six tools
can simulate the digest. They all produce the same peptide list from the
same cleavage rules. The question is what each tool adds beyond the peptide
list: scoring, filtering, ranking, structural context, and programmatic
access.

This vignette is the companion to the [getting-started guide](pepVet-introduction.html). It compares pepVet
against five other tools on a panel of 5 proteins and 5 enzymes.

## Common ground

Three tools (MS-Digest, ExPASy PeptideMass, pepVet) agree closely on the
peptide set from BSA at MC=1.

```{r overlap-table}
knitr::kable(overlap, caption = "BSA + trypsin MC=1 peptide overlap")
```

```{r overlap-fig, eval=has_ggplot2, fig.width=8, fig.height=5, fig.cap="Peptide overlap among MS-Digest, ExPASy PeptideMass, and pepVet for BSA trypsin MC=1. The lighter segment marks peptides outside the 116 shared by all three tools.", fig.alt="Stacked bars showing 154 peptides from pepVet, 130 from MS-Digest, and 151 from ExPASy, with 116 common to all three."}
ggplot2::ggplot(overlap, ggplot2::aes(x = Tool)) +
  ggplot2::geom_col(
    ggplot2::aes(y = N_unique),
    fill = "#AFC6D5", color = "black", width = 0.65
  ) +
  ggplot2::geom_col(
    ggplot2::aes(y = Shared_all_three),
    fill = "#3B7A9E", color = "black", width = 0.65
  ) +
  ggplot2::geom_text(
    ggplot2::aes(y = N_unique, label = N_unique),
    vjust = -0.35
  ) +
  ggplot2::scale_y_continuous(
    limits = c(0, max(overlap$N_unique) * 1.12),
    expand = c(0, 0)
  ) +
  ggplot2::labs(
    title = "Peptide overlap: BSA trypsin MC=1",
    subtitle = "Dark bars show the 116 peptides shared by all three tools",
    x = NULL, y = "Number of peptides"
  ) +
  ggplot2::theme_minimal(base_size = 10) +
  ggplot2::theme(panel.grid.major.x = ggplot2::element_blank())
```

116 peptides are common to all three. pepVet returns the most because it
applies no default mass or length filter. MS-Digest and ExPASy filter
short and hydrophobic fragments by default. Mass correlation between
MS-Digest and ExPASy is 0.999983, confirming consistent residue-mass
physics across all three tools.

## How tools differ

Six tools. They differ in form factor and output type.

```{r cap-table}
knitr::kable(caps, caption = "Tool capability matrix")
```

MS-Digest (Baker, Clauser, and Burlingame, UCSF) returns a tab-delimited
table of peptide masses and coordinates given a protein sequence and an
enzyme. It does not score or rank digests. Its value is mass accuracy.

PeptideMass (Wilkins et al., 1997) accepts a UniProt accession and
returns neutral average masses with pI. It accounts for signal peptides
and known PTMs from UniProt annotations. No scoring or ranking.

Protein Cleaver (Koulouras et al., 2025, Frontiers in Bioinformatics) is
an R Shiny wrapper around `cleaver`. It flags peptides as identifiable or
non-identifiable using configurable length and mass windows. It adds 3D
structure mapping and retention time prediction. Source analysis shows it
uses the same cleavage engine as pepVet but a different mass table
(e.g., Ala 71.079 vs 71.037 Da). No GRAVY filter, no composite score, no
verdict, no presets. Shiny only with heavy dependencies.

ProteaseGuru (Ludwig et al., 2023, Journal of Proteome Research) is a
Windows .NET 8 desktop application. It compares multiple proteases using
SSRCalc hydrophobicity, coverage maps, and shared versus exclusive
peptide tracking. No composite quality score, no presets, no programmatic
access.

PeptideRanger (Rosenberg and Zhang, 2024, Analytical Chemistry) is an R
package that predicts peptide detectability via random forest trained on
CPTAC and ProteomicsDB detection rates. It produces an RF_score (0 to 1)
per peptide. PeptideRanger and pepVet measure orthogonal properties.
pepVet scores digest quality before acquisition (protein level).
PeptideRanger scores detection probability for an existing digest (peptide
level).

## pepVet scoring in practice

pepVet scores 5 proteins across 5 enzymes. The composite score (0 to 1)
combines 5 components: length suitability, sequence coverage, peptide
count, hydrophobicity, and charge.

```{r enzyme-fig, eval=has_ggplot2, fig.width=8, fig.height=5, fig.cap="Enzyme comparison across five proteins. Dashed lines mark the Moderate and Good verdict thresholds. Each bar is one enzyme. pepVet's composite score is a weighted sum of 5 components.", fig.alt="Grouped bar chart of composite scores by protein and enzyme with dashed verdict thresholds."}
ggplot2::ggplot(
  scores,
  ggplot2::aes(x = protein, y = composite, fill = enzyme)
) +
  ggplot2::geom_col(
    position = "dodge", color = "black", alpha = 0.85
  ) +
  ggplot2::geom_hline(
    yintercept = c(0.40, 0.65),
    linetype = "dashed", linewidth = 0.5,
    color = c("#D4A76A", "#3A8C5F")
  ) +
  ggplot2::scale_fill_manual(
    values = enzyme_fill,
    labels = enzyme_labels[names(enzyme_fill)]
  ) +
  ggplot2::scale_y_continuous(
    limits = c(0, 1), expand = c(0, 0.02)
  ) +
  ggplot2::labs(
    title = "Enzyme comparison across proteins",
    y = "Composite score", x = NULL, fill = "Enzyme"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "bottom")
```

Best enzyme per protein:
* BSA (serum albumin): trypsin (0.885, Good).
* H3 (histone): Lys-C (0.769, Good). Histone H3 has few arginines.
* BACE1 (membrane protease): chymotrypsin-high (0.769, Good). BACE1 is
  enriched in hydrophobic domains chymotrypsin handles well.
* Lysozyme C: trypsin (0.825, Good).
* Ubiquitin (fusion): Lys-C (0.885, Good). Polyubiquitin chains have K-G
  linkages Lys-C resolves.

Two pairs score Poor: BACE1 + Lys-C (0.389) and Lysozyme C + Glu-C
(0.269). Both have too few cleavage sites to produce a useful digest.

### Workflow presets

Six presets adjust length range, GRAVY range, and scoring weights. For an
easy protein like BSA, all presets give similar results. For BACE1,
a transmembrane protein, presets that widen the length and GRAVY windows
include more peptides.

```{r preset-fig, eval=has_ggplot2, fig.width=10, fig.height=5, fig.cap="Workflow preset effects on valid peptide count. Grey bars show all digested peptides and coloured bars show peptides accepted by each preset. BSA is comparatively stable, while BACE1 and H3 show greater variation.", fig.alt="Faceted bar chart of total and valid peptide counts across 6 presets for BSA, H3, and BACE1."}
ggplot2::ggplot(presets, ggplot2::aes(x = preset)) +
  ggplot2::geom_col(
    ggplot2::aes(y = n_total),
    fill = "#D8DDE6", alpha = 0.7, width = 0.85
  ) +
  ggplot2::geom_col(
    ggplot2::aes(y = n_valid, fill = verdict),
    color = "black", alpha = 0.85, width = 0.85
  ) +
  ggplot2::geom_text(
    ggplot2::aes(y = n_valid, label = n_valid),
    vjust = -0.3, size = 3.2
  ) +
  ggplot2::facet_wrap(~protein, nrow = 1) +
  ggplot2::scale_fill_manual(values = verdict_fill) +
  ggplot2::labs(
    title = "Preset effects on valid peptide count",
    x = NULL, y = "Number of peptides", fill = "Verdict"
  ) +
  ggplot2::theme_minimal(base_size = 10) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(
      angle = 35, hjust = 1, size = 8
    ),
    legend.position = "bottom",
    strip.background = ggplot2::element_rect(
      fill = "#F0F0F0", color = NA
    ),
    strip.text = ggplot2::element_text(face = "bold")
  )
```

The targeted preset (narrow length 8-20, narrow GRAVY -0.8 to 0.4) keeps
fewest valid peptides. The FFPE preset (wide length 6-30, wider GRAVY -1.0
to 0.8) keeps the most. The verdict for BACE1 changes from Moderate
(standard, 0.573) to Good (ffpe_degraded, 0.655).

## Pipeline with PeptideRanger

pepVet selects an enzyme and identifies valid peptides. PeptideRanger
scores each valid peptide for detection probability. The combined pipeline
covers both digest quality and peptide-level detectability.

Enrichment analysis across 25 protein-enzyme combinations quantifies the
overlap. Enrichment is the difference in mean PR score between
pepVet-valid and pepVet-invalid peptides. Positive enrichment means
pepVet's filter selects higher-detectability peptides.

```{r pr-fig, eval=has_ggplot2, fig.width=9, fig.height=6, fig.cap="PeptideRanger enrichment: mean PR(valid) minus mean PR(invalid). Positive values (green) mean pepVet selects higher-detectability peptides. 18 of 25 combinations show positive enrichment.", fig.alt="Horizontal bar chart of enrichment values for 25 protein-enzyme combinations, colored by sign."}
pr_plot <- pr_data[!is.na(pr_data$enrichment), ]
pr_plot$sign <- ifelse(
  pr_plot$enrichment >= 0, "positive", "negative"
)
pr_plot$enzyme_label <- enzyme_labels[pr_plot$enzyme]
enzyme_order <- tapply(
  pr_plot$enrichment, pr_plot$enzyme_label, mean, na.rm = TRUE
)
pr_plot$enzyme_label <- factor(
  pr_plot$enzyme_label, levels = names(sort(enzyme_order))
)

ggplot2::ggplot(
  pr_plot,
  ggplot2::aes(x = enrichment, y = enzyme_label, fill = sign)
) +
  ggplot2::geom_col(color = "black", alpha = 0.85, width = 0.7) +
  ggplot2::geom_vline(
    xintercept = 0, linewidth = 0.5, color = "grey50"
  ) +
  ggplot2::scale_fill_manual(values = c(
    "positive" = "#3A8C5F", "negative" = "#C46A6A"
  )) +
  ggplot2::facet_grid(protein ~ ., scales = "free_y", space = "free_y") +
  ggplot2::labs(
    title = "PeptideRanger enrichment by pepVet validity filter",
    x = "Enrichment in PR score", y = NULL, fill = NULL
  ) +
  ggplot2::theme_minimal(base_size = 10) +
  ggplot2::theme(
    legend.position = "none",
    panel.grid.major.y = ggplot2::element_blank(),
    strip.text.y = ggplot2::element_text(
      angle = 0, hjust = 1, face = "bold", size = 9
    ),
    strip.background = ggplot2::element_rect(
      fill = "#F0F0F0", color = NA
    )
  )
```

Enrichment is positive in 18 of 25 combinations. Largest effects: BACE1
asp-n (+0.24, membrane domains excluded by GRAVY), Ubiquitin Glu-C (+0.24,
extreme fragments excluded), H3 trypsin (+0.21, short fragments removed).
Negative enrichment occurs in combos with few valid peptides (n < 7) where
the mean is unstable.

A two-stage pipeline run looks like this:

```{r pr-code, eval = requireNamespace("PeptideRanger", quietly = TRUE)}
preset <- pepvet_preset("standard")
ev <- evaluate_digest(
  system.file("extdata", "P02769.fasta", package = "pepVet"),
  enzyme = "trypsin", missed_cleavages = 1L,
  gravy_range = preset$gravy_range,
  length_range = preset$length_range
)
valid <- ev$peptides
valid <- valid[valid$length >= preset$length_range[1] &
               valid$length <= preset$length_range[2], ]
pr_predict <- getExportedValue("PeptideRanger", "peptide_predictions")
pr_model <- getExportedValue("PeptideRanger", "RFmodel_ProteomicsDB")
pr_res <- pr_predict(
  valid$peptide,
  prediction_model = pr_model
)
head(pr_res)
```

## Binary versus graded: pepVet vs Protein Cleaver

Protein Cleaver uses a binary identifiable flag from length and mass.
pepVet uses a graded multi-component score. Simulation of Protein
Cleaver's logic on the same peptides:

```{r pc-table}
knitr::kable(pc_data,
             caption = "Protein Cleaver identifiability vs pepVet validity")
```

```{r pc-fig, eval=has_ggplot2, fig.width=8, fig.height=5, fig.cap="Overlap of pepVet valid peptides with Protein Cleaver identifiable peptides. 'Both' means both tools accept the peptide. 'PC identifiable only' passes the Protein Cleaver filter but not pepVet's GRAVY filter.", fig.alt="Bar chart comparing peptides accepted by both tools with peptides accepted only by Protein Cleaver for BSA, H3, and BACE1."}
pc_plot <- rbind(
  data.frame(
    protein = pc_data$protein,
    classifier = "Both (pepVet + PC)",
    count = pc_data$both
  ),
  data.frame(
    protein = pc_data$protein,
    classifier = "PC identifiable only",
    count = pc_data$pc_only
  )
)

ggplot2::ggplot(
  pc_plot,
  ggplot2::aes(x = protein, y = count, fill = classifier)
) +
  ggplot2::geom_col(
    position = "dodge", color = "black", alpha = 0.85, width = 0.6
  ) +
  ggplot2::geom_text(
    ggplot2::aes(label = count),
    position = ggplot2::position_dodge(width = 0.6),
    vjust = -0.3, size = 3.5
  ) +
  ggplot2::scale_fill_manual(values = c(
    "Both (pepVet + PC)" = "#3A8C5F",
    "PC identifiable only" = "#C46A6A"
  )) +
  ggplot2::labs(
    title = "pepVet vs Protein Cleaver",
    x = NULL, y = "Number of peptides", fill = "Classification"
  ) +
  ggplot2::theme_minimal(base_size = 10) +
  ggplot2::theme(legend.position = "bottom")
```

Every pepVet-valid peptide is also PC-identifiable. PC marks more as
identifiable because it has no GRAVY filter. Protein Cleaver tests length
and mass. pepVet tests coverage, hydrophobicity, charge, and workflow fit.

## When to use what

pepVet alone is the right choice for quick enzyme ranking, method
development with presets, and manuscript-grade digest quality
documentation. Adding PeptideRanger gives peptide-level detection
probabilities for the same digest. Adding Protein Cleaver adds 3D
structural context when a PDB file is available. MS-Digest and ExPASy are
best when only precise mass information is needed. pepVet delivers a
graded, multi-component quality score with workflow presets, verdict
classification, and sensitivity analysis in a single R package.

## Session info

```{r session-info}
sessionInfo()
```
