---
title: "Getting Started with pepVet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with pepVet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(pepVet)
data("aa_properties", package = "pepVet")
```

`pepVet` is a Bioconductor-oriented package for simulating proteolytic
digestion and evaluating peptide sets for proteomics workflows.

The current implementation already includes a working digestion layer,
reference FASTA fixtures, and amino-acid reference data. The scoring and
evaluation layers come later, but the digest API is live and validated.

## Reference data conventions

`pepVet` ships a small amino acid reference table, `aa_properties`, that is
used by the early scoring and validation helpers.

The `molecular_weight` column stores free amino acid monoisotopic masses. If
you need peptide residue masses instead, subtract the mass of water
(18.01056 Da) from those values.

The `pKa_side_chain` column stores conventional reference pKa values for the
ionizable side chains `C`, `D`, `E`, `H`, `K`, `R`, and `Y`. Those values are
appropriate for lightweight package heuristics, but they should not be treated
as context-specific protein pKa predictions.

```{r aa-properties-preview}
aa_properties
```

## Digest proteins

`digest_protein()` accepts several input shapes and always returns a tibble
with `protein_id`, `peptide`, `start`, `end`, `length`, and
`missed_cleavages`.

```{r digest-character}
digest_protein("MKWVTFISLLFLFSSAYSR")
```

```{r digest-aastring}
digest_protein(Biostrings::AAString("MKWVTFISLLFLFSSAYSR"), enzyme = "trypsin")
```

## Supported enzyme names

pepVet validates enzyme names against a pinned registry of cleaver-compatible
rules. A few commonly useful examples are:

- `trypsin`
- `lysc`
- `glutamyl endopeptidase`
- `asp-n endopeptidase`
- `chymotrypsin-high`
- `thermolysin`

The validation layer also trims whitespace and normalizes case, so values such
as `" Trypsin "` are accepted and normalized internally.

## FASTA input

The package ships pinned FASTA fixtures in `inst/extdata/`. You can digest
single-entry or multi-entry FASTA files directly, and the FASTA record names
are preserved in `protein_id`.

```{r digest-fasta}
bsa_path <- system.file("extdata", "P02769.fasta", package = "pepVet")
digest_protein(bsa_path, enzyme = "trypsin")
```

```{r digest-fasta-multi}
isoform_path <- system.file(
  "extdata",
  "P37840_isoforms.fasta",
  package = "pepVet"
)

head(digest_protein(isoform_path, enzyme = "trypsin"))
```

## Missed cleavages

Missed cleavages are expanded explicitly in pepVet so peptide coordinates stay
exact even when peptides overlap.

```{r digest-missed-cleavages}
digest_protein("AKRTPK", enzyme = "trypsin", missed_cleavages = 1L)
```

```{r session-info}
sessionInfo()
```
