---
title: "Getting Started with pepVet"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Getting Started with pepVet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(pepVet)
data("aa_properties", package = "pepVet")
```

pepVet is a Bioconductor-oriented R package for evaluating proteolytic digests
before you spend time at the bench. Given a protein sequence and an enzyme, it
tells you how suitable the resulting peptides are for LC-MS/MS detection.
The package walks a single conceptual pipeline — from raw sequence all the way
to a ranked enzyme recommendation — and every step is individually accessible.

## The core pipeline

```
Protein sequence
      │
      ▼
digest_protein()    →  tibble of peptides (one row per peptide)
      │
      ▼
score_peptides()    →  one-row tibble of component scores + verdict
      │
      ▼
evaluate_digest()   →  convenience wrapper: digest + score + params in one call
      │
      ▼
compare_digests()   →  run evaluate_digest across N enzymes, rank by composite
      │
      ▼
recommend_enzyme()  →  return the name of the top-ranked enzyme
      │
      ▼
batch_evaluate()    →  run evaluate_digest across all proteins in a FASTA
      │
      ▼
digest_report()     →  styled console output for any of the above
```

## Protein input

Every pepVet function that takes a protein sequence accepts the same flexible
input shapes. You can pass:

- a raw character sequence
- a named character vector (one entry per protein)
- a `Biostrings::AAString` or `AAStringSet`
- a path to a FASTA file (single- or multi-entry)

The package ships pinned FASTA fixtures in `inst/extdata/` that are used
throughout this vignette.

```{r paths}
bsa_path <- system.file("extdata", "P02769.fasta", package = "pepVet")
h3_path  <- system.file("extdata", "P68431.fasta", package = "pepVet")
proteome_path <- system.file(
  "extdata", "small_proteome_50_proteins.fasta", package = "pepVet"
)
```

## Step 1 — Digest a protein

`digest_protein()` cleaves a protein sequence using any cleaver-compatible
enzyme rule and returns a tibble with one row per peptide fragment.

```{r digest-basic}
digest_protein(bsa_path, enzyme = "trypsin", missed_cleavages = 1L)
```

The output columns are:

| Column | Description |
|---|---|
| `protein_id` | Full FASTA header or supplied name |
| `peptide` | Amino acid sequence of the fragment |
| `start` | 1-based start position in the parent protein |
| `end` | 1-based end position (inclusive) |
| `length` | Peptide length in residues |
| `missed_cleavages` | Number of internal cleavage sites skipped |

Missed cleavages are expanded explicitly inside pepVet so that start/end
coordinates remain exact and every combination of adjacent peptides is
represented.

```{r digest-mc}
# mc=0: strict cleavage only
digest_protein("AKRTPK", enzyme = "trypsin", missed_cleavages = 0L)

# mc=1: also include once-skipped joins
digest_protein("AKRTPK", enzyme = "trypsin", missed_cleavages = 1L)
```

Enzyme names are validated against a registry of 40 cleaver-compatible rules.
Common examples: `trypsin`, `lysc`, `glutamyl endopeptidase`,
`asp-n endopeptidase`, `chymotrypsin-high`, `thermolysin`. The validation
layer normalises case and trims whitespace, so `" Trypsin "` is accepted.

## Step 2 — Score the peptide set

`score_peptides()` summarises a digest tibble into per-protein component scores
and a weighted composite suitability verdict.

```{r score-basic}
digest_result <- digest_protein(bsa_path, enzyme = "trypsin",
                                missed_cleavages = 1L)
score_peptides(digest_result)
```

**What counts as a "valid" peptide?** Only peptides between 7 and 25 residues
are counted for detection-window scoring (S_length, S_coverage, S_count).
Fragments outside that window still appear in the digest tibble — they are
simply excluded from the numerator of those components.

The five standard components:

| Score | Numerator / logic | Optimal direction |
|---|---|---|
| `S_length` | Valid peptides / total peptides | Higher → more peptides in the LC-MS window |
| `S_coverage` | Non-overlapping valid residues / protein length | Higher → less dark proteome |
| `S_count` | Soft-normalised valid peptide count | Higher → more IDs per protein |
| `S_hydro` | GRAVY-based mean hydrophobicity of valid peptides | Near 0 → balanced retention |
| `S_charge` | Valid peptides with ≥ 1 K/R/H / total valid | Higher → better ESI ionisation |

Composite score verdict thresholds: **Good** ≥ 0.70 · **Moderate** ≥ 0.40 · **Poor** < 0.40.

### Proteome-aware uniqueness scoring

When you supply a `proteome` digest, `score_peptides()` adds a sixth component
`S_unique` — the fraction of valid peptides that appear in exactly one protein
in the proteome. This matters for protein-level inference: shared peptides can
be attributed to multiple proteins and inflate or confuse quantification.

```{r score-proteome, eval=FALSE}
# Digest the proteome background first
proteome_digest <- digest_protein(proteome_path, enzyme = "trypsin")

# Score BSA in the context of that proteome
bsa_digest <- digest_protein(bsa_path, enzyme = "trypsin")
score_peptides(bsa_digest, proteome = proteome_digest)
```

Proteome-aware scoring also switches the default weight set to give `S_unique`
a 20 % share: `c(0.20, 0.20, 0.15, 0.15, 0.10, 0.20)`.

## Step 3 — Evaluate in one call

`evaluate_digest()` wraps `digest_protein()` + `score_peptides()` into a
single call and returns a named list.

```{r evaluate}
ev <- evaluate_digest(bsa_path, enzyme = "trypsin", missed_cleavages = 1L)
names(ev)
```

```{r evaluate-scores}
ev$scores
```

```{r evaluate-params}
ev$params
```

The `$params` element records the resolved enzyme name, missed cleavages, and
the full protein IDs found in the input — useful for logging or downstream
annotation.

## Step 4 — Compare enzymes

`compare_digests()` runs `evaluate_digest()` across a vector of enzyme names
for a single protein and returns a tibble sorted from best to worst composite
score.

```{r compare}
comp <- compare_digests(
  bsa_path,
  enzymes = c("trypsin", "lysc", "glutamyl endopeptidase",
              "asp-n endopeptidase", "chymotrypsin-high"),
  missed_cleavages = 1L
)
comp
```

`compare_digests()` rejects multi-protein input by design — enzyme comparison
is a protein-level decision and mixing proteins would produce a meaningless
aggregate score.

## Step 5 — Get a recommendation

`recommend_enzyme()` runs `compare_digests()` internally and returns the name
of the best-scoring enzyme as a character string. If multiple enzymes tie on
the composite score, all tied names are returned in alphabetical order.

```{r recommend}
recommend_enzyme(
  bsa_path,
  enzymes = c("trypsin", "lysc", "glutamyl endopeptidase",
              "asp-n endopeptidase", "chymotrypsin-high"),
  missed_cleavages = 1L
)
```

## Step 6 — Batch across a proteome

`batch_evaluate()` runs `evaluate_digest()` independently for every protein in
a multi-FASTA file and returns a named list. Each element is identical to the
list you would get from calling `evaluate_digest()` directly on that single
protein.

```{r batch, eval=FALSE}
batch <- batch_evaluate(proteome_path, enzyme = "trypsin",
                        missed_cleavages = 1L)

# How many proteins were evaluated?
length(batch)

# Score of the first protein
batch[[1]]$scores
```

Because each protein is evaluated independently, `batch_evaluate()` parallelises
safely — wrap it in `BiocParallel::bplapply()` if you are processing thousands
of proteins.

## Step 7 — Report to the console

`digest_report()` produces a styled, colour-coded summary for either an
`evaluate_digest()` result (single protein, bar-chart style) or a
`compare_digests()` tibble (ranked table style). It returns its input
invisibly so pipeline chains continue uninterrupted.

```{r report-single}
digest_report(ev)
```

```{r report-compare}
digest_report(comp)
```

Because `digest_report()` returns `invisible(x)`, you can insert it anywhere
in a pipeline to print a human-readable checkpoint without breaking the chain:

```{r pipe, eval=FALSE}
# The printed report appears; the tibble flows on to recommend_enzyme
winner <- compare_digests(bsa_path,
    enzymes = c("trypsin", "lysc", "glutamyl endopeptidase")) |>
  digest_report() |>
  recommend_enzyme()
```

## A challenging protein — Histone H3.1

Histone H3.1 is a useful sanity check because trypsin performs poorly on it.
The protein has a highly basic N-terminal tail rich in lysine and arginine,
so trypsin over-digests it into dozens of very short fragments that fall below
the 7-residue detection threshold.

```{r h3-trypsin}
ev_h3_trypsin <- evaluate_digest(h3_path, enzyme = "trypsin")
ev_h3_trypsin$scores$verdict
```

LysC recognises only lysine and cleaves less frequently on H3.1, producing
longer peptides that score better overall:

```{r h3-compare}
compare_digests(
  h3_path,
  enzymes = c("trypsin", "lysc", "glutamyl endopeptidase",
              "asp-n endopeptidase")
)
```

This is the kind of non-obvious result pepVet is designed to surface before
you run a gel.

## Amino acid reference data

The package ships `aa_properties`, a 20-row tibble with physical and chemical
properties for the standard amino acids used by the scoring engine.

```{r aa-properties}
aa_properties
```

The `kyte_doolittle` column contains the Kyte-Doolittle (1982) hydrophobicity
scale used in `S_hydro`. The `pKa_side_chain` column holds reference pKa
values for the seven ionisable residues (C, D, E, H, K, R, Y) and is used in
`S_charge` classification.

## Session info

```{r session-info}
sessionInfo()
```
