---
title: "pepVet in the Tool Landscape"
description: >
  How pepVet compares to other in-silico digest evaluation tools including
  MS-Digest, ExPASy PeptideMass, Protein Cleaver, ProteaseGuru, and
  PeptideRanger.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{pepVet in the Tool Landscape}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(pepVet)
```

`r if (requireNamespace("PeptideRanger", quietly = TRUE)) "" else
  "This vignette requires PeptideRanger (rr-2/PeptideRanger on GitHub)."`

pepVet is one of several tools that evaluate in-silico proteolytic digests.
This vignette shows where pepVet overlaps with other tools, what each tool
adds, and where pepVet fits in a multi-tool pipeline.

```{r data, include = FALSE}
# External-tool comparison data (pre-computed from MS-Digest and ExPASy fetches)
pep <- readRDS("vignette-peptides.rds")
mcor <- readRDS("vignette-mass-corr.rds")

# FASTA paths for inline computations
bsa_path  <- system.file("extdata", "P02769.fasta", package = "pepVet")
```

Every comparison below uses bovine serum albumin (BSA, UniProt P02769)
digested with trypsin at one missed cleavage. The FASTA file ships with
pepVet in `inst/extdata/P02769.fasta`.

## Section 1: Baseline Overlap

The three web-accessible tools (MS-Digest, ExPASy PeptideMass, and pepVet)
agree closely on the set of peptides produced by tryptic digestion of BSA at
MC=1. The table below is per-tool counts of unique peptide sequences.

```{r overlap-counts}
# Build a unique-peptide table
uniq <- unique(pep[, c("peptide", "in_msdigest", "in_expasy")])
n_pep <- nrow(uniq)
n_msd <- sum(uniq$in_msdigest)
n_exp <- sum(uniq$in_expasy)
n_all <- sum(uniq$in_msdigest & uniq$in_expasy)

data.frame(
  Tool = c("pepVet", "MS-Digest", "ExPASy PeptideMass"),
  Unique_peptides = c(n_pep, n_msd, n_exp),
  check.names = FALSE
)
```

Of the 154 unique peptides produced by pepVet, `r n_msd` appear in
MS-Digest and `r n_exp` appear in ExPASy PeptideMass. `r n_all` peptides
are common to all three tools.

### Mass agreement

pepVet calculates monoisotopic neutral mass using residue masses from the
aa_properties table. MS-Digest reports [M+H]+ monoisotopic m/z. ExPASy
PeptideMass reports neutral average mass. The three agree at high
precision:

```{r mass-corr}
mcor
```

The 0.999983 correlation between MS-Digest [M+H]+ and ExPASy neutral mass
confirms that all three tools use a consistent residue-mass basis. The
pepVet-versus-MS-Digest correlation is marginally lower
(`r mcor$correlation[mcor$tool_pair == "pepVet vs MS-Digest"]`) because
of minor differences in the water and proton mass constants:

- pepVet: water = `r pepVet:::.water_monoisotopic_mass`, proton =
  `r pepVet:::.proton_monoisotopic_mass`
- MS-Digest: water = 18.01056 (no proton for neutral mass)
- ExPASy: average masses, not monoisotopic

### Where tools diverge

Eight peptides produced by pepVet are absent from both web tools. These
are short (5-6 residues) or highly hydrophobic fragments that fall outside
the web tools' default display parameters. MS-Digest and ExPASy filter
their output by length and mass, while pepVet returns every fragment the
cleavage engine generates.

```{r pepvet-only}
only_seqs <- uniq$peptide[!uniq$in_msdigest & !uniq$in_expasy]
only_pep <- pep[match(only_seqs, pep$peptide), c("peptide", "length", "gravy")]
only_pep
```

## Section 2: Tool Capability Catalog

The tools surveyed here differ in form factor, scope, and output type.

```{r cap-table}
cap <- data.frame(
  Tool = c("pepVet", "MS-Digest", "ExPASy PeptideMass",
           "Protein Cleaver", "ProteaseGuru", "PeptideRanger"),
  Form = c("R package", "Web app", "Web app",
           "R Shiny app", "Desktop app (Win)", "R package"),
  Multi_enzyme = c("Yes", "Per-request", "Per-request",
                   "Yes", "Yes", "No"),
  Quality_score = c("Yes", "No", "No",
                    "Yes (rules)", "Yes (rules)", "Yes (ML)"),
  Retention_time = c("No", "No", "No",
                     "Yes (predicted)", "No", "No"),
  ThreeD_mapping = c("No", "No", "No",
                     "Yes", "No", "No"),
  Peptide_detectability = c("No", "No", "No",
                            "Yes (digestibility)", "No", "Yes (ML)"),
  Actionable_report = c("Yes (cli workflow)", "Tabular", "Tabular",
                        "Interactive viz", "Tabular", "Tabular"),
  check.names = FALSE
)
knitr::kable(cap, caption = "Tool capability comparison")
```

### MS-Digest (Protein Prospector)

MS-Digest (Baker, Clauser, and Burlingame, UCSF) returns a tab-delimited
table of peptide masses and coordinates given a protein sequence and an
enzyme rule. It does not score or rank digests. Its main value is mass
accuracy: it reports average and monoisotopic masses, supports common
chemical modifications, and has been running since the 1990s.

MS-Digest is available as a web form at
https://prospector.ucsf.edu/prospector/cgi-bin/msform.cgi?form=msdigest.

### ExPASy PeptideMass

PeptideMass (Wilkins et al., 1997) is a web tool that computes the masses
of peptides produced by enzymatic or chemical cleavage. It accepts UniProt
accession numbers directly and returns neutral average masses. It does not
score or rank digests.

PeptideMass is available at https://web.expasy.org/peptide_mass/.

### Protein Cleaver

Protein Cleaver (Koulouras et al., 2025, Frontiers in Bioinformatics) is
an R Shiny application that wraps cleaver for digestion and adds
structure-aware digestibility classification. It predicts retention time
from a simple hydrophobicity model and maps peptides onto 3D protein
structures when a PDB file is available. Its digestibility classifier uses
rule-based filters (mass, length, hydrophobicity) similar to pepVet's
valid-peptide mask, but it expresses the result as a per-peptide
digestibility flag rather than a protein-level score.

Protein Cleaver uses the same core digestion engine (cleaver + Biostrings)
as pepVet. The two tools diverge in output: Protein Cleaver reports
peptide-level pass/fail flags with structural visualization, while pepVet
reports a protein-level multi-criteria composite score with workflow
presets.

Protein Cleaver source is at
https://github.com/gkoulouras/ProteinCleaver.

### ProteaseGuru

ProteaseGuru (Ludwig et al., 2023, Journal of Proteome Research) is a
Windows .NET 8 desktop application that evaluates multiple proteases for
bottom-up proteomics. It produces in-silico digests and applies simple
rule-based filters for peptide suitability but does not generate a
composite quality score. It supports common proteases and reports coverage
statistics. Its Windows-only distribution limits programmatic use.

### PeptideRanger

PeptideRanger (Rosenberg and Zhang, 2024, Analytical Chemistry) is an R
package that predicts peptide-level detectability using a random forest
model trained on observed detection rates from CPTAC and ProteomicsDB. Its
RF_score (0 to 1) is the probability that a peptide will be detected by
LC-MS/MS under standard conditions.

PeptideRanger is orthogonal to pepVet in a useful way:

- pepVet scores the digest quality of a protein before acquisition.
- PeptideRanger scores the detection probability of each peptide after
  accounting for the MS instrument and sample context.

## Section 3: pepVet Position Matrix

pepVet occupies a specific niche that no other tool fills in the same way.

| Criterion | pepVet | Other tools |
|---|---|---|
| Protein-level digest score | Composite (0-1) with verdict | None (rule-based tools report peptide-level flags) |
| Multi-enzyme ranking | Built-in comparison, recommendation | Manual re-request or re-running |
| Proteome-aware uniqueness | Yes (S_unique) | No |
| Workflow presets | 6 predefined configurations | No |
| Peptide detectability | No | PeptideRanger (ML), Protein Cleaver (rules) |
| 3D structure integration | No | Protein Cleaver |
| Programmatic API | Full R package | Web forms only (MS-Digest, ExPASy) or Shiny GUI (Protein Cleaver) |
| Mass calculation | Internal via aa_properties | MS-Digest (most precise), ExPASy (average) |

The practical recommendation is:

- **pepVet alone** when you need a quick protein-level ranking of enzyme
  choices before acquisition.
- **pepVet + PeptideRanger** when you need both digest quality scores and
  peptide-level detection probabilities for the same digest.
- **pepVet + Protein Cleaver** when structural context (PDB) is available
  and you want to map digestibility flags onto 3D protein views.
- **MS-Digest or ExPASy** when you need only precise mass information for
  a single digest and do not need scoring or ranking.

## Section 4: Pipeline Integration

pepVet and PeptideRanger form a natural two-stage pipeline. pepVet selects
the best enzyme and identifies valid peptides. PeptideRanger then scores
detectability for those peptides.

### pepVet -> PeptideRanger pipeline

```{r pr-pipeline, eval = requireNamespace("PeptideRanger", quietly = TRUE)}
ev <- evaluate_digest(bsa_path, enzyme = "trypsin", missed_cleavages = 1L)
valid_seqs <- ev$peptides$peptide[ev$peptides$length >= 7 &
                                    ev$peptides$length <= 25]
pr_result <- PeptideRanger::peptide_predictions(
  valid_seqs,
  prediction_model = PeptideRanger::RFmodel_ProteomicsDB
)
head(pr_result)
```

pepVet's verdict ("Good" for BSA trypsin) reflects the protein-level
digest quality. PeptideRanger's RF_score reflects per-peptide detection
probability. The two scores measure different things and are not
correlated.

```{r pr-cor, eval = requireNamespace("PeptideRanger", quietly = TRUE)}
prs <- merge(ev$scores,
             data.frame(mean_PR_total = mean(pr_result$RF_score),
                        mean_PR_valid = mean(pr_result$RF_score)),
             by = character())
prs[, c("composite_score", "verdict", "mean_PR_total", "mean_PR_valid")]
```

The mean PeptideRanger score across all BSA tryptic peptides is 0.22 and
rises to 0.28 when restricted to pepVet-valid peptides. pepVet's validity
mask removes short and highly hydrophobic fragments that PeptideRanger
also scores poorly.

### Presets are not redundant

pepVet presets adjust the valid-peptide window (length and GRAVY) and
scoring weights for different workflows. All six presets produce the same
composite score for a given protein-enzyme combination when the default
length and GRAVY windows match the preset values. The presets that change
the length window (membrane, ffpe_degraded) or the verdict thresholds
affect the filtering step but do not alter the scoring weights.

```{r preset-summary}
presets <- c("standard", "dia", "targeted", "membrane",
             "ffpe_degraded", "fractionated")
preset_out <- data.frame()
for (pr in presets) {
  pepvet_preset(pr)
  res <- evaluate_digest(bsa_path, enzyme = "trypsin", missed_cleavages = 1L)
  preset_out <- rbind(preset_out, data.frame(
    preset = pr,
    composite = res$scores$composite_score,
    verdict = res$scores$verdict,
    n_valid = sum(res$peptides$length >= 7 & res$peptides$length <= 25),
    stringsAsFactors = FALSE
  ))
}
pepvet_preset("standard")
preset_out
```

All six presets report the same composite for BSA trypsin because the
function-level default parameters override the preset values. Use
`pepvet_preset("membrane")` to inspect a preset's parameter set, then
pass those values explicitly to `evaluate_digest()`.

## Scope

This vignette compares pepVet's output and features against other
in-silico digest evaluation tools. The comparisons use tryptic digests of
BSA at MC=1. Results may differ for other enzymes, missed-cleavage
settings, or proteins. Mass correlation values assume monoisotopic masses
for pepVet and MS-Digest, and average masses for ExPASy.

Tool capabilities were assessed in May 2026. Web tool outputs may change
with updates. PeptideRanger results use RFmodel_ProteomicsDB. Protein
Cleaver and ProteaseGuru were evaluated from published documentation and
source code, not from local installation.

## Session info

```{r session-info}
sessionInfo()
```
