---
title: "Getting Started with pepVet"
description: >
  A walkthrough of pepVet's core pipeline from a single protein digest to
  multi-enzyme comparison, batch evaluation, and peptide export.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Getting Started with pepVet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(pepVet)
data("aa_properties", package = "pepVet")
```

pepVet evaluates proteolytic digests before bench work begins. Give it a protein sequence and an enzyme. It returns peptide coordinates, score components, a composite verdict, and workflow-aware parameters that let you inspect the result instead of treating it as a black box. A companion vignette ([tool comparison](pepVet-tool-comparison.html)) compares pepVet against five other digestion tools.

## The core pipeline

```
Protein sequence
      │
      v
    digest_protein()    ->  tibble of peptides (one row per peptide)
      │
      v
    score_peptides()    ->  one-row tibble of component scores + verdict
      │
      v
    evaluate_digest()   ->  convenience wrapper: digest + score + params in one call
      │
      v
    pepvet_check()      ->  evaluate + print report in one interactive call
      │
      v
    compare_digests()   ->  run evaluate_digest across N enzymes, rank by composite
      │
      v
    recommend_enzyme()  ->  return the name of the top-ranked enzyme
      │
      v
    batch_evaluate()    ->  run evaluate_digest across all proteins in a FASTA
      │                     returns a flat tibble: one row per protein
      v
    summarize_batch()   ->  proteome-level verdict distribution and score stats
      │
      v
    triage_proteins()   ->  per-protein action recommendations (proceed / try other enzyme …)
      │
      v
    export_peptide_list()  ->  export valid peptides to Skyline, generic CSV, or FASTA
      │
      v
    digest_report()     ->  styled console output for any of the above
```

Every output is programmatic. `digest_protein()` returns a tibble with one row per peptide. `score_peptides()` returns a tibble with one row per protein. `compare_digests()` returns a tibble with one row per enzyme. `evaluate_digest()` returns a named list with the score tibble, the peptide tibble, and the resolved parameters.

## Protein input

Every pepVet function that takes a protein sequence accepts the same input shapes. You can pass:

- a raw character sequence
- a named character vector (one entry per protein)
- a `Biostrings::AAString` or `AAStringSet`
- a path to a FASTA file (single- or multi-entry)

The package ships pinned FASTA fixtures in `inst/extdata/` used throughout this vignette.

```{r paths}
bsa_path <- system.file("extdata", "P02769.fasta", package = "pepVet")
h3_path <- system.file("extdata", "P68431.fasta", package = "pepVet")
proteome_path <- system.file(
  "extdata", "small_proteome_50_proteins.fasta",
  package = "pepVet"
)
```

## Step 1: Digest a protein

`digest_protein()` cleaves a protein sequence using any cleaver-compatible enzyme rule and returns a tibble with one row per peptide fragment.

```{r digest-basic}
digest_protein(bsa_path, enzyme = "trypsin", missed_cleavages = 1L)
```

The output columns are:

| Column | Description |
|---|---|
| `protein_id` | Full FASTA header or supplied name |
| `peptide` | Amino acid sequence of the fragment |
| `start` | 1-based start position in the parent protein |
| `end` | 1-based end position (inclusive) |
| `length` | Peptide length in residues |
| `missed_cleavages` | Number of internal cleavage sites skipped |

Missed cleavages are expanded explicitly so that start/end coordinates remain exact and every combination of adjacent peptides is represented.

```{r digest-mc}
# mc=0: strict cleavage only
digest_protein("AKRTPK", enzyme = "trypsin", missed_cleavages = 0L)

# mc=1: also include once-skipped joins
digest_protein("AKRTPK", enzyme = "trypsin", missed_cleavages = 1L)
```

Enzyme names are validated against a registry of 40 cleaver-compatible rules. Common examples: `trypsin`, `lysc`, `glutamyl endopeptidase`, `asp-n endopeptidase`, `chymotrypsin-high`, `thermolysin`. The validation layer normalises case and trims whitespace, so `" Trypsin "` is accepted.

## Step 1a: Inspect cleavage efficiency

`annotate_cleavage_sites()` adds a second layer for trypsin-family workflows: not just where cleavage can happen, but which local motifs are more likely to resist full digestion.

```{r cleavage-annotations}
annotate_cleavage_sites(bsa_path, enzyme = "trypsin")
```

pepVet currently labels tryptic sites as:

- `low` for proline-blocked (`KP` / `RP`) and acidic P1' (`KD`, `KE`, `RD`, `RE`) motifs
- `medium` for adjacent-basic sites such as `KK`, `KR`, and `RR`
- `high` for other tryptic sites

These labels are sequence-local only. They do not model structure, PTMs, or extended subsite preferences, so they should be read as digestion-risk annotations rather than abundance predictions.

## Step 2: Score the peptide set

`score_peptides()` summarises a digest tibble into per-protein component scores and a weighted composite verdict.

```{r score-basic}
digest_result <- digest_protein(bsa_path,
  enzyme = "trypsin",
  missed_cleavages = 1L
)
score_peptides(digest_result)
```

Only peptides inside the active `length_range` count as valid. The default window is 7 to 25 residues. Fragments outside that window still appear in the digest tibble, but they do not contribute to the component numerators. See the [scoring model](scoring-model.html) article for the formula of each component.

Composite score verdict thresholds are `Good` at `>= 0.65`, `Moderate` at `>= 0.40`, and `Poor` below `0.40`. These thresholds are heuristic. They are not calibrated probabilities. [REF]

## Scoring model

pepVet is not a peptide detectability predictor. Existing ML-based tools learn peptide-level detection patterns from experimental MS data. pepVet does something narrower and more transparent: it ranks the digest quality of a protein for a chosen enzyme-workflow combination using explicit physicochemical criteria.

- ML detectability tools learn nonlinear feature interactions from observed detections.
- pepVet uses a rule-based multi-criteria decision analysis with linear weights.
- ML detectability tools output per-peptide detection probabilities.
- pepVet outputs per-protein digest-quality rankings.

That distinction matters because pepVet scores are designed to explain why a digest looks promising or weak before acquisition, not to estimate the probability that a specific peptide will be observed.

## Evidence basis

Every major scoring parameter has a different level of support. pepVet documents that support explicitly so users can see what is literature-backed, what is a physical constant, and what remains a heuristic pending later calibration.

| Parameter | Category | Current basis | Phase 4 resolution |
|---|---|---|---|
| GRAVY range `[-1.0, +0.6]` | Literature-informed | C18 RP-LC elution guidance and observed under-detection of highly hydrophobic peptides | Calibrate against PeptideAtlas GRAVY distributions |
| Length range `[7, 25]` | Literature-informed | Common bottom-up LC-MS guidance with conservative tails retained | Calibrate against PeptideAtlas peptide-length distributions |
| Default component weights | Expert prior | Chosen for interpretability and digest-planning emphasis, not fit from data | Fit against observed detection summaries |
| Preset weights | Expert prior | Workflow-specific directional logic from literature, not empirically fitted coefficients | Validate by workflow against public benchmark datasets |
| Verdict thresholds `0.65 / 0.4` | Heuristic | Chosen as readable ranking cutoffs | Calibrate with ROC/logistic analysis |
| Enzyme-class fallback constants | Literature-informed | Approximate peptide-length expectations from proteome-wide enzyme behavior | Validate against observed digest medians |
| Fraction-based scoring form | Design choice | Simple, transparent normalization rather than a learned nonlinear model | Compare against alternative forms |
| pKa set and proton mass | Literature-backed / physical constant | Lehninger-style pKa set and NIST proton mass | No calibration for proton mass; pI validated against ExPASy |

## Known limitations

- Verdict thresholds are heuristic labels, not calibrated classifications.
- Weight vectors are expert priors, not empirically trained coefficients.
- PTMs, chemical labels, and digestion chemistry beyond the current rules are ignored.
- GRAVY and peptide-length windows assume conventional C18 reversed-phase LC with ESI.
- `S_charge` reflects extra internal basic-residue richness, not whether a peptide can ionize at all.
- Proteome-aware uniqueness is only as relevant as the proteome digest you supply.

## Scope

pepVet scores are interpretable rankings, not calibrated probabilities. The composite score has no unit. It ranks proteins by digest quality within a given enzyme-workflow combination.

Cross-workflow comparisons are not valid when the resolved scoring configuration changes. To make that explicit, `score_peptides()` records a `preset_used` column. Named presets are reported when the resolved GRAVY window, peptide-length window, weights, and pI setting exactly match a shipped preset; otherwise the call is marked `"custom"`.

## Step 2a: Apply a workflow preset

Workflow presets give you a starting parameter set for common experiment types. Each preset returns a list with `gravy_range`, `length_range`, and `weights`, so you can pass the result into `evaluate_digest()` or `score_peptides()` directly.

```{r preset-standard}
pepvet_preset("standard")
```

```{r preset-evaluate}
syn_path <- system.file("extdata", "P37840_isoforms.fasta", package = "pepVet")
syn_proteome <- digest_protein(syn_path, enzyme = "trypsin")
targeted_preset <- pepvet_preset("targeted")

do.call(
  evaluate_digest,
  c(list(sequence = syn_path, enzyme = "trypsin", proteome = syn_proteome), targeted_preset)
)$scores
```

Six presets ship with pepVet. Each bundles a length range, a GRAVY window, and scoring weights for a common experiment type. See the [workflow presets](workflow-presets.html) article for the full reference.

Presets with non-zero `S_unique` weights require a proteome digest. pepVet rejects those presets when `proteome` is missing because the uniqueness term would otherwise be invented.

### Proteome-aware uniqueness scoring

When you supply a `proteome` digest, `score_peptides()` adds a sixth component, `S_unique`. It measures the fraction of valid peptides that appear in exactly one protein in the proteome. This matters for protein-level inference: shared peptides can be attributed to multiple proteins and inflate or confuse quantification.

```{r score-proteome, eval=FALSE}
# Digest the proteome background first
proteome_digest <- digest_protein(proteome_path, enzyme = "trypsin")

# Score BSA in the context of that proteome
bsa_digest <- digest_protein(bsa_path, enzyme = "trypsin")
score_peptides(bsa_digest, proteome = proteome_digest)
```

Proteome-aware scoring switches the default weight set to give `S_unique` a 20% share: `c(S_length = 0.160, S_coverage = 0.279, S_count = 0.181, S_hydro = 0.110, S_charge = 0.070, S_unique = 0.200)`.

## Step 3: Evaluate in one call

`evaluate_digest()` wraps `digest_protein()` and `score_peptides()` into a single call and returns a named list.

```{r evaluate}
ev <- evaluate_digest(bsa_path, enzyme = "trypsin", missed_cleavages = 1L)
names(ev)
```

```{r evaluate-scores}
ev$scores
```

```{r evaluate-params}
ev$params
```

The `$params` element records the resolved enzyme name, missed cleavages, and the full protein IDs found in the input.
It also records `preset_used`, which is the simplest guard against comparing scores that came from different workflow assumptions.

When you need peptide-level cleavage-risk context, request it explicitly:

```{r evaluate-cleavage-efficiency}
ev_eff <- evaluate_digest(
  bsa_path,
  enzyme = "trypsin",
  missed_cleavages = 1L,
  include_cleavage_efficiency = TRUE
)

ev_eff$peptides
ev_eff$scores[, c("protein_id", "n_high_efficiency_sites", "n_low_efficiency_sites")]
```

The protein-level counts are informational. They are not score components, but they help flag proteins where missed cleavages are likely to remain prominent even before you inspect individual peptides.

## Step 4: Compare enzymes

`compare_digests()` runs `evaluate_digest()` across a vector of enzyme names for a single protein and returns a tibble sorted from best to worst composite score.

```{r compare}
comp <- compare_digests(
  bsa_path,
  enzymes = c(
    "trypsin", "lysc", "glutamyl endopeptidase",
    "asp-n endopeptidase", "chymotrypsin-high"
  ),
  missed_cleavages = 1L
)
comp
```

`compare_digests()` rejects multi-protein input by design. Enzyme comparison is a protein-level decision, and mixing proteins would produce a meaningless aggregate score.

## Step 5: Get a recommendation

`recommend_enzyme()` runs `compare_digests()` internally and returns the name of the best-scoring enzyme as a character string. If multiple enzymes tie on the composite score, all tied names are returned in alphabetical order.

```{r recommend}
recommend_enzyme(
  bsa_path,
  enzymes = c(
    "trypsin", "lysc", "glutamyl endopeptidase",
    "asp-n endopeptidase", "chymotrypsin-high"
  ),
  missed_cleavages = 1L
)
```

## Step 6: Batch across a proteome

`batch_evaluate()` runs `evaluate_digest()` independently for every protein in a multi-FASTA file and returns a **flat tibble** with one row per protein. Columns include `protein_id`, `protein_length`, all component scores, `composite_score`, `verdict`, count fields, and four sequence-level difficulty flags.

```{r batch, eval=FALSE}
batch <- batch_evaluate(proteome_path,
  enzyme = "trypsin",
  missed_cleavages = 1L
)

# How many proteins were evaluated?
nrow(batch)

# Score and verdict for the first few proteins
batch[, c("protein_id", "composite_score", "verdict")]
```

Each protein is evaluated independently. The returned tibble slots directly into `summarize_batch()` and `triage_proteins()`.

For proteome-aware uniqueness scoring, supply a background proteome digest:

```{r proteome-onboarding, eval=FALSE}
proteome_digest <- digest_protein(proteome_path, enzyme = "trypsin")
batch <- batch_evaluate(proteome_path,
  enzyme = "trypsin",
  proteome = proteome_digest
)
```

## Step 6a: Summarize a batch

`summarize_batch()` computes proteome-level aggregate statistics from the tibble returned by `batch_evaluate()`. It returns a named list with five elements.

```{r summarize-batch, eval=FALSE}
summary <- summarize_batch(batch)

# Verdict distribution (Good / Moderate / Poor and their percentages)
summary$verdict_counts

# Composite score distribution
summary$score_distribution

# Per-component mean scores: lowest value is the weakest dimension
summary$component_summary

# Proteins in the bottom 10% by composite score
summary$problem_proteins

# Moderate / Poor proteins where hydrophobicity or short-protein flags are set:
# likely candidates for switching to a less specific enzyme
summary$enzyme_switch_candidates
```

`enzyme_switch_candidates` is a heuristic derived from sequence-level difficulty flags, not from running alternative enzymes. Use `compare_digests()` to confirm whether a specific alternative enzyme actually improves a flagged protein.

## Step 6b: Triage proteins

`triage_proteins()` appends an `action` column to the batch tibble with a deterministic recommendation for each protein.

```{r triage, eval=FALSE}
triaged <- triage_proteins(batch)

# Count of each action
table(triaged$action)

# Proteins that should be tried with a different enzyme
triaged[
  triaged$action == "try_other_enzyme",
  c("protein_id", "verdict", "composite_score")
]
```

| Action | Meaning |
|---|---|
| `proceed` | Good verdict. No intervention indicated. |
| `consider_alternative` | Moderate verdict with at least one component score below 0.5. A preset change or missed-cleavage adjustment may help. |
| `try_other_enzyme` | Moderate or Poor verdict with hydrophobicity or short-protein flag. An alternative enzyme is the most likely improvement path. |
| `skip` | No valid peptides or low-complexity sequence. No standard enzyme choice is expected to substantially improve the score. |

## Step 6c: Export a peptide list

`export_peptide_list()` filters valid peptides from any `digest_protein()` tibble and writes them in a format compatible with downstream tools.

```{r export-skyline}
peps <- digest_protein(bsa_path, enzyme = "trypsin", missed_cleavages = 1L)

# Skyline transition list: one row per valid peptide per charge state
# Columns: Protein, Peptide Sequence, Precursor Charge, Precursor Mz
export_peptide_list(peps, format = "skyline", charges = 2:3)
```

```{r export-generic}
# Generic annotated table: all peptide columns plus gravy, pI, and valid flag
export_peptide_list(peps, format = "generic")
```

```{r export-fasta}
# FASTA character vector for valid peptides only
# Each header: >protein_id|start-end
head(export_peptide_list(peps, format = "fasta"))
```

Write directly to a file by passing a path to the `file` argument:

```{r export-file, eval=FALSE}
export_peptide_list(peps, format = "skyline", file = "bsa_transitions.csv")
export_peptide_list(peps, format = "fasta", file = "bsa_peptides.fasta")
```

## Step 7: Quick interactive check

`pepvet_check()` combines `evaluate_digest()` and `digest_report()` into a single call. It is the fastest way to inspect a protein interactively: one function call, one styled console report, result returned invisibly for further use.

```{r pepvet-check}
result <- pepvet_check(bsa_path, enzyme = "trypsin")
```

The returned value is the full `evaluate_digest()` result list, so you can pipe it into any downstream step:

```{r pepvet-check-pipe, eval=FALSE}
result <- pepvet_check(bsa_path, enzyme = "trypsin", missed_cleavages = 1L)
result$scores
result$peptides
```

## Step 8: Report to the console

`digest_report()` produces a styled, colour-coded summary for either an `evaluate_digest()` result (single protein, bar-chart style) or a `compare_digests()` tibble (ranked table style). It returns its input invisibly so pipeline chains continue uninterrupted.

```{r report-single}
digest_report(ev)
```

```{r report-compare}
digest_report(comp)
```

`digest_report()` returns `invisible(x)`, so you can insert it anywhere in a pipeline to print a human-readable checkpoint without breaking the chain:

```{r pipe, eval=FALSE}
# Compare enzymes and print a report
comp <- compare_digests(bsa_path,
  enzymes = c("trypsin", "lysc", "glutamyl endopeptidase")
)
digest_report(comp)

# Get the top recommendation directly from the sequence
winner <- recommend_enzyme(bsa_path,
  enzymes = c("trypsin", "lysc", "glutamyl endopeptidase")
)
```

## A challenging protein: Histone H3.1

Histone H3.1 is a useful sanity check because trypsin performs poorly on it. The protein has a highly basic N-terminal tail rich in lysine and arginine. Trypsin over-digests it into dozens of very short fragments that fall below the 7-residue detection threshold.

```{r h3-trypsin}
ev_h3_trypsin <- evaluate_digest(h3_path, enzyme = "trypsin")
ev_h3_trypsin$scores$verdict
```

LysC recognises only lysine and cleaves less frequently on H3.1, producing longer peptides that score better overall:

```{r h3-compare}
compare_digests(
  h3_path,
  enzymes = c(
    "trypsin", "lysc", "glutamyl endopeptidase",
    "asp-n endopeptidase"
  )
)
```

## Amino acid reference data

The package ships `aa_properties`, a 20-row tibble with physical and chemical properties for the standard amino acids used by the scoring engine.

```{r aa-properties}
aa_properties
```

The `kyte_doolittle` column contains the Kyte-Doolittle (1982) hydrophobicity scale used in `S_hydro`. The `is_basic` flag supports `S_charge`, and the `pKa_side_chain` column holds reference pKa values for the seven ionisable residues (C, D, E, H, K, R, Y) used in `calculate_pI()`.

## How pepVet differs from common tools

pepVet fills a narrow part of the workflow. It is not a spectral library builder and it is not a peptide detectability predictor. It is a decision-support layer for enzyme choice and digest quality before acquisition. [REF]

The [tool-comparison vignette](pepVet-tool-comparison.html) compares pepVet against five tools on a panel of five proteins and five enzymes. It covers baseline peptide overlap, capability differences across 16 dimensions, the scoring model in practice, workflow preset effects, pipeline integration with PeptideRanger, and a binary versus graded classification comparison with Protein Cleaver.

## Where to go next

The remaining articles on this site go deeper into specific topics.

| Article | What it covers |
|---------|---------------|
| [scoring model](scoring-model.html) | Formula and logic for each of the five component scores |
| [enzyme selection](enzyme-selection.html) | Worked comparisons across multiple proteins with a decision table |
| [workflow presets](workflow-presets.html) | The six shipped presets and how to modify them |
| [visualisation](visualisation.html) | All plot functions with examples |
| [weight sensitivity](weight-sensitivity.html) | Dirichlet Monte Carlo analysis of weight robustness |
| [score diagnostics](score-diagnostics.html) | VIF, PCA, and ablation analysis of scoring model structure |
| [tool comparison](pepVet-tool-comparison.html) | Detailed comparison against MS-Digest, ExPASy, Protein Cleaver, ProteaseGuru, and PeptideRanger |

## Session info

```{r session-info}
sessionInfo()
```
