---
title: "Getting Started with pepVet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with pepVet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(pepVet)
data("aa_properties", package = "pepVet")
```

`pepVet` is being built as a Bioconductor-oriented package for simulating
proteolytic digestion and evaluating peptide sets for proteomics workflows.

The current repository state provides the package scaffold only. The primary
public entry point is reserved below and will be implemented in the next phase.

## Reference data conventions

`pepVet` ships a small amino acid reference table, `aa_properties`, that is
used by the early scoring and validation helpers.

The `molecular_weight` column stores free amino acid monoisotopic masses. If
you need peptide residue masses instead, subtract the mass of water
(18.01056 Da) from those values.

The `pKa_side_chain` column stores conventional reference pKa values for the
ionizable side chains `C`, `D`, `E`, `H`, `K`, `R`, and `Y`. Those values are
appropriate for lightweight package heuristics, but they should not be treated
as context-specific protein pKa predictions.

```{r aa-properties-preview}
aa_properties
```

```{r digest-stub}
try(digest_protein("MKWVTFISLLFLFSSAYSR"), silent = TRUE)
```

```{r session-info}
sessionInfo()
```
