---
title: "Getting Started with pepVet"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Getting Started with pepVet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(pepVet)
data("aa_properties", package = "pepVet")
```

pepVet evaluates proteolytic digests before bench work begins. Give it a protein sequence and an enzyme. It returns peptide coordinates, score components, a composite verdict, and workflow-aware parameters that let you inspect the result instead of treating it as a black box.

## The core pipeline

```
Protein sequence
      │
      v
    digest_protein()    ->  tibble of peptides (one row per peptide)
      │
      v
    score_peptides()    ->  one-row tibble of component scores + verdict
      │
      v
    evaluate_digest()   ->  convenience wrapper: digest + score + params in one call
      │
      v
    compare_digests()   ->  run evaluate_digest across N enzymes, rank by composite
      │
      v
    recommend_enzyme()  ->  return the name of the top-ranked enzyme
      │
      v
    batch_evaluate()    ->  run evaluate_digest across all proteins in a FASTA
      │
      v
    digest_report()     ->  styled console output for any of the above
```

Every output is programmatic. `digest_protein()` returns a tibble with one row per peptide. `score_peptides()` returns a tibble with one row per protein. `compare_digests()` returns a tibble with one row per enzyme. `evaluate_digest()` returns a named list with the score tibble, the peptide tibble, and the resolved parameters.

## Protein input

Every pepVet function that takes a protein sequence accepts the same input shapes. You can pass:

- a raw character sequence
- a named character vector (one entry per protein)
- a `Biostrings::AAString` or `AAStringSet`
- a path to a FASTA file (single- or multi-entry)

The package ships pinned FASTA fixtures in `inst/extdata/` used throughout this vignette.

```{r paths}
bsa_path <- system.file("extdata", "P02769.fasta", package = "pepVet")
h3_path  <- system.file("extdata", "P68431.fasta", package = "pepVet")
proteome_path <- system.file(
  "extdata", "small_proteome_50_proteins.fasta", package = "pepVet"
)
```

## Step 1: Digest a protein

`digest_protein()` cleaves a protein sequence using any cleaver-compatible enzyme rule and returns a tibble with one row per peptide fragment.

```{r digest-basic}
digest_protein(bsa_path, enzyme = "trypsin", missed_cleavages = 1L)
```

The output columns are:

| Column | Description |
|---|---|
| `protein_id` | Full FASTA header or supplied name |
| `peptide` | Amino acid sequence of the fragment |
| `start` | 1-based start position in the parent protein |
| `end` | 1-based end position (inclusive) |
| `length` | Peptide length in residues |
| `missed_cleavages` | Number of internal cleavage sites skipped |

Missed cleavages are expanded explicitly so that start/end coordinates remain exact and every combination of adjacent peptides is represented.

```{r digest-mc}
# mc=0: strict cleavage only
digest_protein("AKRTPK", enzyme = "trypsin", missed_cleavages = 0L)

# mc=1: also include once-skipped joins
digest_protein("AKRTPK", enzyme = "trypsin", missed_cleavages = 1L)
```

Enzyme names are validated against a registry of 40 cleaver-compatible rules. Common examples: `trypsin`, `lysc`, `glutamyl endopeptidase`, `asp-n endopeptidase`, `chymotrypsin-high`, `thermolysin`. The validation layer normalises case and trims whitespace, so `" Trypsin "` is accepted.

## Step 2: Score the peptide set

`score_peptides()` summarises a digest tibble into per-protein component scores and a weighted composite verdict.

```{r score-basic}
digest_result <- digest_protein(bsa_path, enzyme = "trypsin",
                                missed_cleavages = 1L)
score_peptides(digest_result)
```

Only peptides inside the active `length_range` count as valid. The default window is 7 to 25 residues. Fragments outside that window still appear in the digest tibble, but they do not contribute to the component numerators.

The five standard components:

| Score | Numerator / logic | Optimal direction |
|---|---|---|
| `S_length` | Valid peptides / total peptides | Higher → more peptides in the LC-MS window |
| `S_coverage` | Non-overlapping valid residues / protein length | Higher → less dark proteome |
| `S_count` | Soft-normalised valid peptide count | Higher → more IDs per protein |
| `S_hydro` | GRAVY-based mean hydrophobicity of valid peptides | Near 0 → balanced retention |
| `S_charge` | Valid peptides with ≥ 1 non-terminal K/R/H / total valid | Higher → more multi-charge potential and richer fragment series |

`v0.0.4` also reports `median_peptide_length`, which is the digest-level denominator used in `S_count`. That value matters when you compare enzymes with very different cleavage densities.

Composite score verdict thresholds are `Good` at `>= 0.70`, `Moderate` at `>= 0.40`, and `Poor` below `0.40`. These thresholds are heuristic. They are not calibrated probabilities. [REF]

## Step 2a: Apply a workflow preset

Workflow presets give you a starting parameter set for common experiment types. Each preset returns a list with `gravy_range`, `length_range`, and `weights`, so you can pass the result into `evaluate_digest()` or `score_peptides()` directly.

```{r preset-standard}
pepvet_preset("standard")
```

```{r preset-evaluate}
syn_path <- system.file("extdata", "P37840_isoforms.fasta", package = "pepVet")
syn_proteome <- digest_protein(syn_path, enzyme = "trypsin")
targeted_preset <- pepvet_preset("targeted")

do.call(
  evaluate_digest,
  c(list(sequence = syn_path, enzyme = "trypsin", proteome = syn_proteome), targeted_preset)
)$scores
```

| Preset | Best fit | Main scoring shift |
|---|---|---|
| `standard` | Routine DDA | Package defaults |
| `dia` | DIA and SWATH | Higher coverage weight |
| `targeted` | SRM, PRM, MRM | Higher uniqueness weight |
| `membrane` | Hydrophobic proteins | Wider GRAVY window |
| `ffpe_degraded` | Degraded material | Wider peptide-length window |
| `fractionated` | SCX or high-pH RP planning | Standard score with fractionation-oriented interpretation |

Presets with non-zero `S_unique` weights require a proteome digest. pepVet rejects those presets when `proteome` is missing because the uniqueness term would otherwise be invented.

### Proteome-aware uniqueness scoring

When you supply a `proteome` digest, `score_peptides()` adds a sixth component, `S_unique`. It measures the fraction of valid peptides that appear in exactly one protein in the proteome. This matters for protein-level inference: shared peptides can be attributed to multiple proteins and inflate or confuse quantification.

```{r score-proteome, eval=FALSE}
# Digest the proteome background first
proteome_digest <- digest_protein(proteome_path, enzyme = "trypsin")

# Score BSA in the context of that proteome
bsa_digest <- digest_protein(bsa_path, enzyme = "trypsin")
score_peptides(bsa_digest, proteome = proteome_digest)
```

Proteome-aware scoring switches the default weight set to give `S_unique` a 20% share: `c(0.20, 0.20, 0.15, 0.15, 0.10, 0.20)`.

## Step 3: Evaluate in one call

`evaluate_digest()` wraps `digest_protein()` and `score_peptides()` into a single call and returns a named list.

```{r evaluate}
ev <- evaluate_digest(bsa_path, enzyme = "trypsin", missed_cleavages = 1L)
names(ev)
```

```{r evaluate-scores}
ev$scores
```

```{r evaluate-params}
ev$params
```

The `$params` element records the resolved enzyme name, missed cleavages, and the full protein IDs found in the input.

## Step 4: Compare enzymes

`compare_digests()` runs `evaluate_digest()` across a vector of enzyme names for a single protein and returns a tibble sorted from best to worst composite score.

```{r compare}
comp <- compare_digests(
  bsa_path,
  enzymes = c("trypsin", "lysc", "glutamyl endopeptidase",
              "asp-n endopeptidase", "chymotrypsin-high"),
  missed_cleavages = 1L
)
comp
```

`compare_digests()` rejects multi-protein input by design. Enzyme comparison is a protein-level decision, and mixing proteins would produce a meaningless aggregate score.

## Step 5: Get a recommendation

`recommend_enzyme()` runs `compare_digests()` internally and returns the name of the best-scoring enzyme as a character string. If multiple enzymes tie on the composite score, all tied names are returned in alphabetical order.

```{r recommend}
recommend_enzyme(
  bsa_path,
  enzymes = c("trypsin", "lysc", "glutamyl endopeptidase",
              "asp-n endopeptidase", "chymotrypsin-high"),
  missed_cleavages = 1L
)
```

## Step 6: Batch across a proteome

`batch_evaluate()` runs `evaluate_digest()` independently for every protein in a multi-FASTA file and returns a named list. Each element matches what you would get from calling `evaluate_digest()` directly on that protein.

```{r batch, eval=FALSE}
batch <- batch_evaluate(proteome_path, enzyme = "trypsin",
                        missed_cleavages = 1L)

# How many proteins were evaluated?
length(batch)

# Score of the first protein
batch[[1]]$scores
```

Each protein is evaluated independently. Wrap in `BiocParallel::bplapply()` if you are processing thousands of proteins.

The setup is short when you want a proteome-aware workflow.

```{r proteome-onboarding, eval=FALSE}
proteome_digest <- digest_protein(proteome_path, enzyme = "trypsin")
batch <- batch_evaluate(proteome_path, enzyme = "trypsin", proteome = proteome_digest)
compare_digests(bsa_path, enzymes = c("trypsin", "lysc"), proteome = proteome_digest)
```

## Step 7: Report to the console

`digest_report()` produces a styled, colour-coded summary for either an `evaluate_digest()` result (single protein, bar-chart style) or a `compare_digests()` tibble (ranked table style). It returns its input invisibly so pipeline chains continue uninterrupted.

```{r report-single}
digest_report(ev)
```

```{r report-compare}
digest_report(comp)
```

`digest_report()` returns `invisible(x)`, so you can insert it anywhere in a pipeline to print a human-readable checkpoint without breaking the chain:

```{r pipe, eval=FALSE}
# The printed report appears; the tibble flows on to recommend_enzyme
winner <- compare_digests(bsa_path,
    enzymes = c("trypsin", "lysc", "glutamyl endopeptidase")) |>
  digest_report() |>
  recommend_enzyme()
```

## A challenging protein: Histone H3.1

Histone H3.1 is a useful sanity check because trypsin performs poorly on it. The protein has a highly basic N-terminal tail rich in lysine and arginine. Trypsin over-digests it into dozens of very short fragments that fall below the 7-residue detection threshold.

```{r h3-trypsin}
ev_h3_trypsin <- evaluate_digest(h3_path, enzyme = "trypsin")
ev_h3_trypsin$scores$verdict
```

LysC recognises only lysine and cleaves less frequently on H3.1, producing longer peptides that score better overall:

```{r h3-compare}
compare_digests(
  h3_path,
  enzymes = c("trypsin", "lysc", "glutamyl endopeptidase",
              "asp-n endopeptidase")
)
```

## Amino acid reference data

The package ships `aa_properties`, a 20-row tibble with physical and chemical properties for the standard amino acids used by the scoring engine.

```{r aa-properties}
aa_properties
```

The `kyte_doolittle` column contains the Kyte-Doolittle (1982) hydrophobicity scale used in `S_hydro`. The `is_basic` flag supports `S_charge`, and the `pKa_side_chain` column holds reference pKa values for the seven ionisable residues (C, D, E, H, K, R, Y) used in `calculate_pI()`.

## How pepVet differs from common tools

pepVet fills a narrow part of the workflow. It is not a spectral library builder and it is not a peptide detectability predictor. It is a decision-support layer for enzyme choice and digest quality before acquisition. [REF]

| Tool | Form | API | Multi-enzyme ranking | Proteome-aware uniqueness |
|---|---|---|---|---|
| Protein Prospector | Web app | No | Manual | No |
| Skyline | Desktop GUI | Limited scripting | Limited | Library-driven |
| cleaver | R package | Yes | No | No |
| pepVet | R package | Yes | Yes | Yes |

## Session info

```{r session-info}
sessionInfo()
```
