---
title: "Getting Started with pepVet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with pepVet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(pepVet)
data("aa_properties", package = "pepVet")
```

`pepVet` is a Bioconductor-oriented package for simulating proteolytic
digestion and evaluating peptide sets for proteomics workflows.

The current implementation already includes a working digestion layer,
reference FASTA fixtures, amino-acid reference data, and a protein-level
scoring engine. The higher-level evaluation and comparison layers come later,
but the digest and scoring APIs are live and validated.

## Reference data conventions

`pepVet` ships a small amino acid reference table, `aa_properties`, that is
used by the early scoring and validation helpers.

The `molecular_weight` column stores free amino acid monoisotopic masses. If
you need peptide residue masses instead, subtract the mass of water
(18.01056 Da) from those values.

The `pKa_side_chain` column stores conventional reference pKa values for the
ionizable side chains `C`, `D`, `E`, `H`, `K`, `R`, and `Y`. Those values are
appropriate for lightweight package heuristics, but they should not be treated
as context-specific protein pKa predictions.

```{r aa-properties-preview}
aa_properties
```

## Digest proteins

`digest_protein()` accepts several input shapes and always returns a tibble
with `protein_id`, `peptide`, `start`, `end`, `length`, and
`missed_cleavages`.

```{r digest-character}
digest_protein("MKWVTFISLLFLFSSAYSR")
```

```{r digest-aastring}
digest_protein(Biostrings::AAString("MKWVTFISLLFLFSSAYSR"), enzyme = "trypsin")
```

## Supported enzyme names

pepVet validates enzyme names against a pinned registry of cleaver-compatible
rules. A few commonly useful examples are:

- `trypsin`
- `lysc`
- `glutamyl endopeptidase`
- `asp-n endopeptidase`
- `chymotrypsin-high`
- `thermolysin`

The validation layer also trims whitespace and normalizes case, so values such
as `" Trypsin "` are accepted and normalized internally.

## FASTA input

The package ships pinned FASTA fixtures in `inst/extdata/`. You can digest
single-entry or multi-entry FASTA files directly, and the FASTA record names
are preserved in `protein_id`.

```{r digest-fasta}
bsa_path <- system.file("extdata", "P02769.fasta", package = "pepVet")
digest_protein(bsa_path, enzyme = "trypsin")
```

```{r digest-fasta-multi}
isoform_path <- system.file(
  "extdata",
  "P37840_isoforms.fasta",
  package = "pepVet"
)

head(digest_protein(isoform_path, enzyme = "trypsin"))
```

## Missed cleavages

Missed cleavages are expanded explicitly in pepVet so peptide coordinates stay
exact even when peptides overlap.

```{r digest-missed-cleavages}
digest_protein("AKRTPK", enzyme = "trypsin", missed_cleavages = 1L)
```

## Score peptide sets

`score_peptides()` summarizes a digest tibble into component scores and a
weighted composite verdict.

```{r score-basic}
digest_result <- digest_protein(bsa_path, enzyme = "trypsin")
score_peptides(digest_result)
```

The default protein-only scoring components are:

- `S_length`
- `S_coverage`
- `S_count`
- `S_hydro`
- `S_charge`

The composite score is a weighted sum of those components, and the verdict is
classified as `Good`, `Moderate`, or `Poor`.

## Proteome-aware uniqueness scoring

If you supply a proteome digest, pepVet adds `S_unique` and uses the
proteome-aware default weight set.

```{r score-proteome-aware}
proteome_digest <- digest_protein(
  c(target = "AAAAAAARAAAAAAAK", background = "AAAAAAARGGGGGGGK"),
  enzyme = "trypsin"
)

score_peptides(
  proteome_digest[proteome_digest$protein_id == "target", ],
  proteome = proteome_digest
)
```

```{r session-info}
sessionInfo()
```
