---
title: "Choosing a Proteolytic Enzyme with pepVet"
description: >
  A practical guide to enzyme selection for LC-MS/MS proteomics. Covers the
  biology of each scoring component, worked multi-enzyme comparisons on real
  proteins, and guidance for difficult targets like histones, membrane proteins,
  and intrinsically disordered proteins.
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(pepVet)
bsa_path   <- system.file("extdata", "P02769.fasta", package = "pepVet")
h3_path    <- system.file("extdata", "P68431.fasta", package = "pepVet")
syn_path   <- system.file("extdata", "P37840_isoforms.fasta", package = "pepVet")
```

## Why enzyme choice matters

Every bottom-up proteomics experiment begins with proteolytic digestion. The enzyme you choose determines the size, sequence coverage, charge-state distribution, and uniqueness of the peptides that reach the mass spectrometer. A poor choice renders an otherwise detectable protein invisible:

- **Too many short peptides.** Fragments shorter than ~7 residues are lost in the void volume of most reversed-phase columns and fail to produce reliable MS/MS spectra.
- **Too few long peptides.** Peptides longer than ~25 residues drift in retention time, produce hard-to-assign fragment ions, and carry poor charge states under standard ESI conditions.
- **Highly hydrophobic peptides.** They stick to sample preparation surfaces and column frits. Yields are variable and reproducibility drops.
- **Peptides without basic residues.** They ionise poorly in positive-mode ESI and produce weak fragmentation patterns.

pepVet quantifies each of these dimensions as a component score, combines them into a weighted composite, and returns a verdict that lets you compare enzymes on equal footing.

## The 40 supported enzymes

pepVet validates against all 40 cleaver-compatible enzyme names. The most relevant for standard bottom-up proteomics are:

| Enzyme | Cleavage rule | Common use case |
|---|---|---|
| `trypsin` | C-terminal to K, R (not before P) | Default choice; ~85 % of published proteomics |
| `lysc` | C-terminal to K only | Histones, basic proteins; longer peptides than trypsin |
| `glutamyl endopeptidase` | C-terminal to E (GluC in phosphate buffer) | Proteins poor in K/R; good for acidic regions |
| `asp-n endopeptidase` | N-terminal to D, C | Phosphoproteomics; orthogonal to trypsin |
| `chymotrypsin-high` | C-terminal to F, Y, W, L (high specificity) | Hydrophobic transmembrane regions |
| `thermolysin` | N-terminal to hydrophobics (L, I, M, V, A, F) | Membrane proteins; orthogonal fragmentation |
| `lysarginase` | C-terminal to K and R | Like trypsin but includes all K and R without Pro exception |
| `pepsin` | C-terminal to F, L; various others | HDX-MS; activity at low pH |

For a complete list, see `?digest_protein` or call:

```{r enzyme-list, eval=FALSE}
# pepVet validates against this registry internally
# You can inspect it by triggering a validation error intentionally:
tryCatch(
  digest_protein("MKWVTFISLLFLFSSAYSR", enzyme = "unknown"),
  error = function(e) message(conditionMessage(e))
)
```

## Worked example 1: BSA (a tractable protein)

BSA (UniProt P02769) is the canonical positive control for proteomics sample preparation. It is a monomeric, soluble, modestly hydrophobic protein with ~580 residues. Trypsin should score well, but let us verify and compare.

```{r bsa-compare}
enzymes_standard <- c(
  "trypsin", "lysc", "glutamyl endopeptidase",
  "asp-n endopeptidase", "chymotrypsin-high"
)

comp_bsa <- compare_digests(bsa_path,
  enzymes = enzymes_standard,
  missed_cleavages = 1L
)
comp_bsa
```

```{r bsa-report}
digest_report(comp_bsa)
```

**Reading the table:** The `>` marker indicates the top-scoring enzyme. Columns are abbreviated: `S_len` = S_length, `S_cov` = S_coverage, `S_cnt` = S_count, `S_hyd` = S_hydro, `S_chg` = S_charge.

**What to look for:**

- `S_coverage` near 1.0 is ideal. Values below 0.6 mean large parts of the protein sequence will go undetected.
- `S_length` below 0.5 signals an enzyme that either over-digests or under-digests the target.
- A `Good` composite (> 0.70) means reliable, reproducible quantification with standard LC-MS/MS parameters.

At mc=1, glutamyl endopeptidase edges ahead of trypsin on BSA because it generates a higher fraction of detectable peptides in the optimal length range. Trypsin is not always the best choice, even for soluble proteins.

### The effect of missed cleavages

Missed cleavages change the score landscape. With mc=0, trypsin over-cuts BSA because many tryptic peptides are adjacent K/R dipeptides that produce sub-7-residue fragments.

```{r bsa-mc}
# mc=0: strict cleavage
comp_mc0 <- compare_digests(bsa_path, enzymes = c("trypsin", "lysc"),
                            missed_cleavages = 0L)
comp_mc0[, c("enzyme", "composite_score", "verdict", "S_length")]

# mc=1: one allowed skip
comp_mc1 <- compare_digests(bsa_path, enzymes = c("trypsin", "lysc"),
                            missed_cleavages = 1L)
comp_mc1[, c("enzyme", "composite_score", "verdict", "S_length")]
```

For most trypsin experiments, mc=1 is the biologically realistic setting because a fraction of cleavage sites is always missed in practice. mc=2 or higher is relevant only for long or poorly digesting peptides.

## Worked example 2: Histone H3.1 (a basic, difficult protein)

Histone H3.1 (UniProt P68431) has a highly basic N-terminal tail rich in K and R. Trypsin cuts after nearly every residue in that tail, producing dozens of 1-3 residue fragments that fall far below the detection window. H3.1 is a classic case where enzyme choice is non-trivial.

```{r h3-compare}
comp_h3 <- compare_digests(h3_path,
  enzymes = c(
    "trypsin", "lysc", "glutamyl endopeptidase", "asp-n endopeptidase"
  )
)
digest_report(comp_h3)
```

LysC cleaves after K only, skipping the many R residues in the tail. The result is longer peptides and substantially better S_length and S_coverage scores. Even scoring only Moderate, LysC improves over trypsin enough to make a meaningful experimental difference.

For basic proteins (histones, protamines, some kinases), always benchmark LysC alongside trypsin. For the most challenging targets, GluC or Asp-N provide orthogonal coverage of basic regions that both trypsin and LysC miss.

## Worked example 3: Alpha-synuclein isoforms (proteome-aware scoring)

Alpha-synuclein (UniProt P37840) is an intrinsically disordered protein associated with Parkinson's disease. The reference file ships three isoforms (canonical plus two alternatively spliced variants). When you need to distinguish between isoforms, peptide uniqueness is critical. A peptide shared between two isoforms cannot tell you which one it came from.

```{r syn-proteome-aware}
# First build the proteome background from all three isoforms
syn_proteome <- digest_protein(syn_path, enzyme = "trypsin")

# Score isoform 1 against that background
syn_main <- system.file("extdata", "P37840_isoforms.fasta", package = "pepVet")
ev_syn <- evaluate_digest(syn_main, enzyme = "trypsin",
                          proteome = syn_proteome)
ev_syn$scores
```

The `S_unique` value quantifies what fraction of valid peptides are unique to one isoform within the background. A low `S_unique` means most peptides are shared, and isoform-level quantification will be unreliable regardless of how well the other scores look.

## Choosing enzymes for special cases

### Membrane proteins

Transmembrane helices are rich in hydrophobic residues (L, I, V, F, M, W). Trypsin rarely cuts within a helix because K and R are excluded from the membrane interior by the positive-inside rule. Large transmembrane stretches go uncovered.

Recommended alternatives:

- `chymotrypsin-high`: cuts C-terminal to F, Y, W, L; enters hydrophobic segments directly.
- `thermolysin`: cuts N-terminal to hydrophobics; thermostable; active in detergent conditions compatible with membrane protein solubilisation.
- `glutamyl endopeptidase`: cuts C-terminal to E; E residues flank helix ends in polar loop regions.

A two-enzyme strategy is common for membrane proteins: trypsin for soluble loops and chymotrypsin or thermolysin for transmembrane helices.

### Phosphoproteomics

Sequence coverage around phosphorylation sites requires enzymes whose cleavage windows overlap the site of interest.

- `asp-n endopeptidase`: cuts N-terminal to D; complementary windows to trypsin around serine/threonine-rich regions.
- `glutamyl endopeptidase`: cuts after E; particularly useful for pY sites near acidic clusters.
- `lysc` + trypsin sequential digestion: longer initial LysC peptides are secondarily cut by trypsin; effective for basic, tightly phosphorylated sequences.

### Large intrinsically disordered proteins

IDPs like huntingtin, FUS, or TDP-43 contain long low-complexity regions with very few K/R. Standard trypsin digestion leaves stretches of hundreds of residues uncovered.

Recommended: `glutamyl endopeptidase` (cuts after E, which appears throughout most low-complexity regions) or `chymotrypsin-high` (cuts after aromatic/hydrophobic residues, enriched in aggregation-prone segments).

## Using `recommend_enzyme()` in scripts

For scripted workflows, `recommend_enzyme()` returns a character string you can feed directly into downstream code.

```{r recommend}
# Returns the single best enzyme name
recommend_enzyme(bsa_path,
  enzymes = enzymes_standard,
  missed_cleavages = 1L
)
```

```{r recommend-h3}
# H3.1 correctly identifies lysc over trypsin
recommend_enzyme(h3_path,
  enzymes = c("trypsin", "lysc", "glutamyl endopeptidase")
)
```

In automated pipelines you might iterate over a target list:

```{r recommend-loop, eval=FALSE}
targets <- c(bsa_path, h3_path)

for (path in targets) {
  winner <- recommend_enzyme(path,
    enzymes = c("trypsin", "lysc", "glutamyl endopeptidase"))
  message(basename(path), "  →  ", winner)
}
```

## Summary decision tree

```
Is the protein basic (many K/R)?
  Yes → Try lysc first; compare with glutamyl endopeptidase
  No  → Does the protein have transmembrane domains?
          Yes → Try chymotrypsin-high or thermolysin; two-enzyme strategy
          No  → Try trypsin at mc=1 first
                  composite > 0.7 (Good)? → Done
                  No → compare_digests() across all candidates
```

Use `digest_report()` to review the score breakdown at any stage. The bar-chart view from `evaluate_digest()` reveals which sub-score is dragging the composite down and points to the specific enzyme property you need to address.

```{r session-info}
sessionInfo()
```
