---
title: "Choosing a Proteolytic Enzyme with pepVet"
description: >
  A practical guide to enzyme selection for LC-MS/MS proteomics with score interpretation and worked examples for basic, membrane, and isoform-aware proteins.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Choosing a Proteolytic Enzyme with pepVet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(pepVet)
bsa_path <- system.file("extdata", "P02769.fasta", package = "pepVet")
h3_path <- system.file("extdata", "P68431.fasta", package = "pepVet")
syn_path <- system.file("extdata", "P37840_isoforms.fasta", package = "pepVet")
```

## Why enzyme choice is not a formality

Enzyme selection affects which peptides are available to a bottom-up LC-MS/MS workflow. Experimental comparisons have found that using different proteases can change protein identifications and sequence coverage, although the result depends on the sample and method (Swaney et al. 2010).

- It can cut often and produce many peptides below pepVet's active length window.
- It can cut rarely and leave large regions in long peptides.
- It can shift the digest toward peptide properties that are less suitable for the intended workflow.

pepVet represents those tradeoffs as score components for direct comparison. It does not predict identification. It ranks enzyme options under an explicit set of digest-level assumptions.

## A fast comparison workflow

A direct workflow uses a side-by-side enzyme comparison.

```{r bsa-compare}
enzymes_standard <- c(
  "trypsin",
  "lysc",
  "glutamyl endopeptidase",
  "asp-n endopeptidase",
  "chymotrypsin-high"
)

comp_bsa <- compare_digests(
  bsa_path,
  enzymes = enzymes_standard,
  missed_cleavages = 1L
)

comp_bsa
```

```{r bsa-report}
digest_report(comp_bsa)
```

The comparison table reports each score component alongside the composite.

- `S_length` shows how much of the digest lands inside the active peptide-length window.
- `S_coverage` reports the fraction of the parent protein covered by valid peptides.
- `S_count` uses an enzyme-aware expected peptide length, so trypsin and GluC are not forced into the same denominator.
- `S_hydro` and `S_charge` report the active GRAVY-window fraction and internal basic-residue fraction.

When two enzymes have similar composite scores, inspect the components. One enzyme can have higher coverage while another has a higher length- or GRAVY-window fraction.

## Worked examples

Three fixed examples show how the same comparison workflow behaves for a reference protein, a basic protein, and a small isoform set.

### Worked example 1: BSA

BSA provides a fixed example for comparing several enzyme rules under one scoring configuration.

```{r bsa-slim}
comp_bsa[, c("enzyme", "S_length", "S_coverage", "S_count", "composite_score", "verdict")]
```

The BSA readout does not identify a universally correct enzyme. It shows how enzyme choice changes peptide density and the package's suitability components under the active settings. In the current release, `S_count` uses the observed digest structure instead of a fixed tryptic expectation.

#### Cleavage efficiency is not uniform

pepVet also reports sequence-local cleavage-efficiency classes for supported trypsin-family rules.

```{r bsa-cleavage-efficiency}
bsa_sites <- annotate_cleavage_sites(bsa_path, enzyme = "trypsin")
bsa_sites[bsa_sites$position %in% c(140L, 155L, 167L), ]
```

On the shipped BSA fixture, pepVet labels the KP motif at position 140 as `low` and the adjacent-basic motifs KK at 155 and RR at 167 as `medium`. The labels come from local motif rules. They are not cleavage probabilities.

The current annotation rules use local P1-P1' sequence context only. They do not model structural accessibility or PTMs and do not replace the digest score.

#### Missed cleavages change the ranking

```{r bsa-missed-cleavages}
comp_mc0 <- compare_digests(
  bsa_path,
  enzymes = c("trypsin", "lysc"),
  missed_cleavages = 0L
)

comp_mc1 <- compare_digests(
  bsa_path,
  enzymes = c("trypsin", "lysc"),
  missed_cleavages = 1L
)

comp_mc0[, c("enzyme", "composite_score", "S_length", "median_peptide_length")]
comp_mc1[, c("enzyme", "composite_score", "S_length", "median_peptide_length")]
```

The two tables show how allowing one missed cleavage changes peptide length and the component scores for this BSA example. They do not estimate an experimental missed-cleavage distribution.

### Worked example 2: Histone H3.1

Histone H3.1 is a basic protein with a lysine- and arginine-rich N-terminal tail. Trypsin often over-cuts this class of target.

```{r h3-compare}
comp_h3 <- compare_digests(
  h3_path,
  enzymes = c("trypsin", "lysc", "glutamyl endopeptidase", "asp-n endopeptidase")
)

comp_h3
```

```{r h3-report}
digest_report(comp_h3)
```

In this in-silico example, trypsin has weak `S_length` and `S_coverage`, while LysC and the acidic-side enzymes have higher values for one or both components. This result is specific to the H3.1 sequence and the active pepVet settings. It does not establish a preferred enzyme for histone experiments.

Histone workflows often require special digestion strategies because lysine- and arginine-rich sequences can yield many short tryptic products. One established bottom-up method blocks lysine residues before trypsin digestion so that cleavage occurs at arginine (Garcia et al. 2007). Middle-down methods retain longer histone-tail analytes and address a different experimental question (Sidoli et al. 2015).

### Worked example 3: Isoform-aware scoring

Alpha-synuclein isoforms are a compact example of when uniqueness matters as much as length and coverage.

```{r syn-proteome}
syn_proteome <- digest_protein(syn_path, enzyme = "trypsin")

ev_syn <- evaluate_digest(
  syn_path,
  enzyme = "trypsin",
  proteome = syn_proteome,
  weights = pepvet_preset("targeted")$weights
)

ev_syn$scores
```

For isoform-level work, inspect `S_unique` against a relevant comparison proteome rather than relying on a protein-only composite score.

## A practical reading guide

The table maps component patterns to follow-up checks.

| Pattern | Likely issue | Next action |
|---|---|---|
| Low `S_length`, low `S_count` | Few products inside the active length window | Inspect the length distribution, then compare a missed-cleavage setting or another enzyme |
| Low `S_coverage`, moderate `S_count` | Valid peptides leave uncovered regions | Compare an enzyme with different cleavage specificity |
| Strong `S_length`, weak `S_hydro` | Many valid-length peptides fall outside the active GRAVY window | Inspect the GRAVY distribution before changing the window |
| Strong core scores, weak `S_unique` | Shared peptides across homologs or isoforms | Supply a proteome digest and use a uniqueness-aware preset |

## Special cases

### Membrane proteins

For a membrane target, inspect whether hydrophobic regions remain uncovered under the selected enzyme. Compare other cleavage rules and the `membrane` preset only when their resolved assumptions match the workflow.

### Degraded material

The `ffpe_degraded` preset includes peptides from 6 to 30 residues. Use it or a custom range only when that inclusion rule matches the workflow, and record the resolved configuration.

### Targeted assays

The `targeted` preset assigns more weight to `S_unique` and requires a relevant proteome digest. Inspect that component alongside the composite score.

## Using recommend_enzyme in scripts

```{r recommend}
recommend_enzyme(
  h3_path,
  enzymes = c("trypsin", "lysc", "glutamyl endopeptidase")
)
```

`recommend_enzyme()` returns the top enzyme name. If scores tie within floating-point tolerance, it returns all tied names. Scripts can use the return value, but it contains no component-level explanation.

## References

Garcia BA, Mollah S, Ueberheide BM, et al. Chemical derivatization of histones for facilitated analysis by mass spectrometry. *Nature Protocols*. 2007;2:933-938. doi: [10.1038/nprot.2007.106](https://doi.org/10.1038/nprot.2007.106).

Sidoli S, Lin S, Karch KR, Garcia BA. Bottom-Up and Middle-Down Proteomics Have Comparable Accuracies in Defining Histone Post-Translational Modification Relative Abundance and Stoichiometry. *Analytical Chemistry*. 2015;87(6):3129-3133. doi: [10.1021/acs.analchem.5b00072](https://doi.org/10.1021/acs.analchem.5b00072).

Swaney DL, Wenger CD, Coon JJ. Value of Using Multiple Proteases for Large-Scale Mass Spectrometry-Based Proteomics. *Journal of Proteome Research*. 2010;9(3):1323-1329. doi: [10.1021/pr900863u](https://doi.org/10.1021/pr900863u).

## Session info

```{r session-info}
sessionInfo()
```
