---
title: "Choosing a Proteolytic Enzyme with pepVet"
description: >
  A practical guide to enzyme selection for LC-MS/MS proteomics. Covers score
  interpretation, worked comparisons, and targeted advice for basic proteins,
  membrane proteins, and isoform-aware workflows.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %VignetteIndexEntry{Choosing a Proteolytic Enzyme with pepVet}
  %VignetteEngine{knitr::rmarkdown}
  %VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(pepVet)
bsa_path <- system.file("extdata", "P02769.fasta", package = "pepVet")
h3_path <- system.file("extdata", "P68431.fasta", package = "pepVet")
syn_path <- system.file("extdata", "P37840_isoforms.fasta", package = "pepVet")
```

## Why enzyme choice is not a formality

Enzyme selection decides whether a protein yields peptides that fit routine LC-MS/MS. A poor enzyme can fail in several ways.

- It can cut too often and flood the digest with peptides below the detection window.
- It can cut too rarely and leave large regions in peptides that are too long to identify reproducibly.
- It can bias the digest toward peptides that retain poorly, ionize poorly, or fail to separate isoforms.

pepVet converts those tradeoffs into score components you can compare directly. The goal is not to predict identification with certainty. The goal is to rank enzyme options using transparent digest-level evidence. [REF]

## A fast comparison workflow

The shortest useful workflow is a side-by-side enzyme comparison.

```{r bsa-compare}
enzymes_standard <- c(
  "trypsin",
  "lysc",
  "glutamyl endopeptidase",
  "asp-n endopeptidase",
  "chymotrypsin-high"
)

comp_bsa <- compare_digests(
  bsa_path,
  enzymes = enzymes_standard,
  missed_cleavages = 1L
)

comp_bsa
```

```{r bsa-report}
digest_report(comp_bsa)
```

The comparison table is compact, but it exposes the scoring logic.

- `S_length` shows how much of the digest lands inside the active peptide-length window.
- `S_coverage` shows how much of the parent protein is covered by valid peptides.
- `S_count` uses an enzyme-aware expected peptide length, so trypsin and GluC are not forced into the same denominator.
- `S_hydro` and `S_charge` show whether the surviving peptides stay in the LC window and retain extra internal basic-residue richness for stronger multi-charge fragmentation.

When two enzymes are close on the composite score, inspect the components before deciding. One enzyme may win on coverage while another wins on peptide quality. That difference matters when the experiment is targeted or proteome-aware.

## Worked example 1: BSA

BSA is a stable positive-control protein. It is useful because most common enzymes produce interpretable digests, but they do not produce the same digest.

```{r bsa-slim}
comp_bsa[, c("enzyme", "S_length", "S_coverage", "S_count", "composite_score", "verdict")]
```

The main readout on BSA is not that one enzyme is universally correct. The main readout is that enzyme choice changes both the density and the detectability of the peptide set. The current release makes that clearer because `S_count` is tied to the observed digest structure instead of a fixed tryptic expectation.

### Cleavage efficiency is not uniform

pepVet `0.1.0` also exposes a second question that matters in practice: not every theoretical cleavage site is equally likely to cut cleanly.

```{r bsa-cleavage-efficiency}
bsa_sites <- annotate_cleavage_sites(bsa_path, enzyme = "trypsin")
bsa_sites[bsa_sites$position %in% c(140L, 155L, 167L), ]
```

On the shipped BSA fixture, pepVet flags low-efficiency proline-blocked sites such as the KP motif at position 140, and medium-efficiency adjacent-basic sites such as KK at 155 and RR at 167. That does not mean those cuts fail completely. It means they are more likely to contribute to missed-cleavage products than an isolated high-efficiency tryptic site.

Use these annotations to explain digest behavior, not to replace the score. The current rules are deliberately narrow: local P1-P1' sequence context only, with no structural accessibility or PTM modeling.

### Missed cleavages change the ranking

```{r bsa-missed-cleavages}
comp_mc0 <- compare_digests(
  bsa_path,
  enzymes = c("trypsin", "lysc"),
  missed_cleavages = 0L
)

comp_mc1 <- compare_digests(
  bsa_path,
  enzymes = c("trypsin", "lysc"),
  missed_cleavages = 1L
)

comp_mc0[, c("enzyme", "composite_score", "S_length", "median_peptide_length")]
comp_mc1[, c("enzyme", "composite_score", "S_length", "median_peptide_length")]
```

That shift is expected. Strict cleavage produces many short tryptic fragments. Allowing one missed cleavage often moves the digest toward a more realistic experimental distribution.

## Worked example 2: Histone H3.1

Histone H3.1 is a basic protein with a lysine- and arginine-rich N-terminal tail. Trypsin often over-cuts this class of target.

```{r h3-compare}
comp_h3 <- compare_digests(
  h3_path,
  enzymes = c("trypsin", "lysc", "glutamyl endopeptidase", "asp-n endopeptidase")
)

comp_h3
```

```{r h3-report}
digest_report(comp_h3)
```

The relevant pattern is usually a weak `S_length` and `S_coverage` for trypsin, followed by improvement under LysC or an orthogonal acidic-side enzyme. That is why histones are a standard counterexample to the assumption that trypsin is always the best first pass. [REF]

## Worked example 3: Isoform-aware scoring

Alpha-synuclein isoforms are a compact example of when uniqueness matters as much as length and coverage.

```{r syn-proteome}
syn_proteome <- digest_protein(syn_path, enzyme = "trypsin")

ev_syn <- evaluate_digest(
  syn_path,
  enzyme = "trypsin",
  proteome = syn_proteome,
  weights = pepvet_preset("targeted")$weights
)

ev_syn$scores
```

If `S_unique` is weak, a high composite score without a proteome background can still be misleading for isoform-level work. That is why presets with non-zero uniqueness weights should be used only when you have a relevant comparison proteome.

## A practical reading guide

Use the score pattern to decide what to try next.

| Pattern | Likely issue | Next action |
|---|---|---|
| Low `S_length`, low `S_count` | Over-digestion into short peptides | Allow one missed cleavage or compare against LysC |
| Low `S_coverage`, moderate `S_count` | Cleavage leaves dark regions | Try a more orthogonal enzyme such as GluC or Asp-N |
| Strong `S_length`, weak `S_hydro` | Peptides are detectable in size but poor for LC | Widen GRAVY expectations only if the biology justifies it |
| Strong core scores, weak `S_unique` | Shared peptides across homologs or isoforms | Supply a proteome digest and use a uniqueness-aware preset |

## Special cases

### Membrane proteins

Trypsin often avoids the transmembrane core. That hurts `S_coverage` even when the soluble loops digest cleanly. Compare trypsin with `chymotrypsin-high` or `thermolysin`, and consider the `membrane` preset if hydrophobic peptides are expected rather than treated as outliers.

### Degraded material

For FFPE or otherwise degraded material, the standard 7 to 25 residue window can be too strict. Use `pepvet_preset("ffpe_degraded")` or pass a wider `length_range` explicitly. The change should be documented because it changes what pepVet counts as a valid peptide.

### Targeted assays

In SRM, PRM, and MRM workflows, you may care less about broad coverage and more about stable, unique peptides. Use the `targeted` preset and inspect `S_unique` before trusting the composite score.

## Using recommend_enzyme in scripts

```{r recommend}
recommend_enzyme(
  h3_path,
  enzymes = c("trypsin", "lysc", "glutamyl endopeptidase")
)
```

`recommend_enzyme()` returns the top enzyme name, or several names if the top scores are tied within floating-point tolerance. That makes it safe for scripted triage, but you should still inspect the comparison table when the experiment has constraints beyond the default scoring model.

## Summary

pepVet is most useful when enzyme choice is uncertain, not when it is obvious. Start with `compare_digests()`, inspect the component pattern, then move to a preset or custom parameter set only when the experiment justifies it.

```{r session-info}
sessionInfo()
```
