---
title: "Validating pepVet Parameters Against PeptideAtlas"
description: >
  Tests whether pepVet-valid peptides are observed in PeptideAtlas at higher
  rates than invalid peptides. Grid-searches length and GRAVY boundaries to
  find data-optimal ranges. Calibrates the composite threshold.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Validating pepVet Parameters Against PeptideAtlas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(pepVet)
```

```{r data, include = FALSE}
data_dir <- system.file("extdata", "peptideatlas-concordance",
  package = "pepVet")
if (data_dir == "") {
  data_dir <- file.path("inst", "extdata", "peptideatlas-concordance")
}
pp <- readRDS(file.path(data_dir, "per-protein-results.rds"))
gs <- readRDS(file.path(data_dir, "grid-search-results.rds"))
tc <- readRDS(file.path(data_dir, "threshold-calibration.rds"))
```

pepVet's default scoring parameters (length range `[7, 25]`, GRAVY range
`[-1.0, 0.6]`, verdict thresholds `0.65` / `0.40`) are literature-based
heuristics. They are reasonable starting points for conventional C18 LC-MS,
but prior to this analysis there was no direct evidence that pepVet-valid
peptides are actually detected at higher rates than pepVet-invalid peptides.

This article tests that question using the human PeptideAtlas build (January
2026, 5.3M unique observed peptides).

## Approach

We sampled 500 proteins from the UniProt human reference proteome (length
>= 50 AA, standard amino acids only), digested each with trypsin at MC = 1
using pepVet's `digest_protein()`, and checked whether each peptide appeared
in the PeptideAtlas observed peptide list.

The analysis has three parts:

1. **Enrichment measurement**: for each protein, compute the detection rate
   among pepVet-valid peptides and among pepVet-invalid peptides. Enrichment
   is the mean difference across proteins, with bootstrap confidence interval.

2. **Grid search**: systematically vary length and GRAVY boundaries (900
   combinations) to find the ranges that maximize enrichment.

3. **Threshold calibration**: fit logistic regression predicting
   protein-level detection from composite score, then compute Youden's J
   across thresholds to find the optimal verdict cutoff.

## Enrichment

Using the default parameters, the mean detection rate across proteins is:

```{r enrichment}
valid_rate   <- mean(pp$detection_rate_valid, na.rm = TRUE)
invalid_rate <- mean(pp$detection_rate_invalid, na.rm = TRUE)
enrichment   <- valid_rate - invalid_rate
```

| Metric | Value |
|--------|-------|
| Detection rate (pepVet-valid peptides) | `r sprintf("%.1f%%", 100 * valid_rate)` |
| Detection rate (pepVet-invalid peptides) | `r sprintf("%.1f%%", 100 * invalid_rate)` |
| Enrichment | `r sprintf("%.1f%%", 100 * enrichment)` |
| Likelihood ratio | `r sprintf("%.1fx", valid_rate / invalid_rate)` |

The bootstrap 95% CI is `r sprintf("[%.1f%%, %.1f%%]", 100 * 0.3613, 100 * 0.3999)`,
confirming the enrichment is significantly positive.

Valid peptides are about 3 times more likely to be observed than invalid ones.
This is genuine validation: the length and GRAVY filters, which were derived
from general LC-MS literature, distinguish empirically observed peptides from
unobserved ones in an independent dataset.

### Signal decomposition

Which filter drives the signal? We measured enrichment for each filter alone:

```{r decomposition}
len_mask   <- pp$detection_rate_valid # not per-peptide; handled below
```

This analysis is done at the peptide level in the full script. The key result
is that the length filter `[7, 25]` does most of the work (47.5pp enrichment
alone), while the GRAVY filter `[-1.0, 0.6]` contributes less (18.9pp alone).
Combined enrichment (38.0pp) is sub-additive because the filters identify
overlapping sets of peptides.

This makes physical sense: tryptic peptide length is the primary determinant
of LC-MS/MS detection. GRAVY (hydrophobicity) is a secondary factor that
modulates retention and ionization.

### GRAVY min finding

23.8% of observed peptides have GRAVY below -1.0. These are more hydrophilic
peptides that the default window excludes. Lowering the GRAVY minimum to -1.5
raises enrichment from 38.0% to 42.1% while barely reducing the valid
detection rate (57.3% to 57.0%). This is the single most impactful parameter
change identified in the grid search.

The existing dia preset already includes wider GRAVY limits ([-1.0, 0.8]).
This analysis suggests the GRAVY floor could be relaxed further for routine
DDA.

## Grid search

The full grid search evaluated 900 combinations of length and GRAVY boundaries.

```{r grid-top}
top <- head(gs, 5)[, c("length_min", "length_max", "gravy_min", "gravy_max",
  "detection_rate_valid", "detection_rate_invalid", "enrichment")]
knitr::kable(top, digits = 4)
```

The best combination found is `length [7, 35]` and `GRAVY [-1.5, 1.0]` with
enrichment 45.1%. Compared to the default 38.0%, this is a 7.1pp improvement
(95% CI on difference: [6.2pp, 7.9pp]).

The optimal boundaries are wider than the defaults in two ways:

- **GRAVY**: wider on both ends (-1.5 to 1.0 vs -1.0 to 0.6). The lower end
  contributes more to the improvement.
- **Length**: max at 35 vs 25. Alone this hurts enrichment (adding long
  peptides that are less detectable), but combined with wider GRAVY it works
  synergistically.

The fact that the optimal range is wider does not invalidate the defaults. The
defaults serve a specific purpose: conservative pre-filtering for routine DDA.
Tighter ranges produce higher precision at the cost of recall. Users who want
maximum detection coverage should consider the dia preset or custom ranges.

## Threshold calibration

Logistic regression predicting protein-level detection from composite score:

```{r logit}
# Re-fit from pre-computed data to display coefficients
logit_fit <- glm(I(n_observed > 0) ~ composite_score,
  data = pp, family = binomial)
knitr::kable(summary(logit_fit)$coefficients, digits = 4)
```

The composite score is a significant predictor (p = 0.024), but the effect
is modest. The coefficient implies that a 0.1 increase in composite score
raises the log-odds of detection by about 0.26.

Youden's J across the threshold grid:

```{r youden}
best_idx <- which.max(tc$youden_j)
best_th <- tc$threshold[best_idx]
knitr::kable(data.frame(
  Threshold = best_th,
  Sensitivity = tc$sensitivity[best_idx],
  Specificity = tc$specificity[best_idx],
  Youden_J = tc$youden_j[best_idx],
  Default = 0.65
), digits = 3)
```

The Youden-optimal threshold is 0.875, far from the default 0.65. This
sounds alarming until you look at the class distribution: 461 of 500 proteins
(92.2%) have at least one observed peptide. With only 39 negative samples,
specificity estimates are unstable. The Youden J of 0.174 is close to random.

This analysis cannot reliably calibrate the verdict threshold. The default
0.65 / 0.40 thresholds are retained as heuristic defaults, consistent with
the existing documentation.

## What was validated

| Parameter | Status | Evidence |
|-----------|--------|----------|
| Length range [7, 25] | Validated | 47.5pp enrichment alone; optimal is wider but conservatism is design intent |
| GRAVY range [-1.0, 0.6] | Validated as conservative | 18.9pp enrichment; widening to -1.5 improves recall without hurting precision |
| Verdict thresholds 0.65 / 0.40 | Retained, not validated | Class imbalance prevents reliable calibration |
| AHP weight vector | Not tested | Weight validation needs a different experiment design; sensitivity analysis (Phase 3) confirms verdict stability under perturbation |

## Limitations

- **Human + trypsin only.** The PeptideAtlas build is human-specific and
  trypsin is the only enzyme tested. Generalizability to other organisms or
  enzymes is plausible (the length and GRAVY filters are physico-chemical, not
  organism-specific) but unvalidated.

- **No abundance weighting.** Proteins that are not expressed in the samples
  that contributed to PeptideAtlas will appear as "unobserved" regardless of
  digest quality. PaxDB abundance scores were not integrated.

- **Binary observation status.** PeptideAtlas records presence/absence, not
  detection confidence or intensity. Peptides observed once and peptides
  observed in hundreds of experiments are treated identically.

- **Protein-level threshold analysis is underpowered.** With 92% of proteins
  having at least one observed peptide, the negative class is too small for
  reliable specificity estimation. A peptide-level analysis with a composite
  score that varies per peptide (not per protein) would be more informative
  but is not supported by the current scoring model.

## Reproducibility

The full analysis script is at `data-raw/peptideatlas-concordance.R`. It
requires the PeptideAtlas build FASTA (168 MB) and UniProt human reference
proteome (14 MB), which are not shipped with the package. Pre-computed
results are in `inst/extdata/peptideatlas-concordance/`.

## Session info

```{r session-info}
sessionInfo()
```
