---
title: "Understanding the pepVet Scoring Model"
description: >
  A detailed guide to pepVet score components, weight handling, preset logic,
  verdict thresholds, and the practical limits of the current model.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %VignetteIndexEntry{Understanding the pepVet Scoring Model}
  %VignetteEngine{knitr::rmarkdown}
  %VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(pepVet)
bsa_path <- system.file("extdata", "P02769.fasta", package = "pepVet")
h3_path <- system.file("extdata", "P68431.fasta", package = "pepVet")
```

## Overview

`score_peptides()` reduces a digest tibble to one row per protein. That row contains component scores, a composite score, a verdict, and the digest-level `median_peptide_length` used to interpret peptide density.

The model is intentionally simple. It is designed to expose why a digest looks promising or weak, not to hide the reasoning inside a single opaque number.

## Positioning

pepVet is not a peptide detectability predictor. It does not try to estimate whether a specific peptide will be identified from empirical MS training data. Instead, it ranks the overall suitability of a protein digest for a chosen enzyme-workflow combination.

- ML detectability models learn nonlinear interactions from observed peptide detections.
- pepVet uses a rule-based multi-criteria decision analysis with explicit, linear weights.
- ML detectability models output per-peptide probabilities.
- pepVet outputs per-protein digest-quality rankings.

That means pepVet should be interpreted as a transparent decision-support model for pre-acquisition planning, not as a substitute for peptide-level detectability predictors.

## Valid peptides

Most pepVet components operate on valid peptides. A peptide is valid if its length falls inside the active `length_range`.

The default window is 7 to 25 residues.

$$
v(p) = \mathbf{1}\{7 \le |p| \le 25\}
$$

Peptides outside that window still appear in `digest_protein()` output. They are real cleavage products. They are excluded because routine reversed-phase LC-MS/MS usually performs poorly at the extremes. [REF]

```{r boundary-demo}
short_seq <- "AKAKAKAKAKAKAKAKAKAKAKAKAKAKAKAKAK"
evaluate_digest(short_seq, enzyme = "trypsin")$scores
```

## Component scores

### S_length

`S_length` is the fraction of peptides that are valid.

$$
S_{length} = \frac{N_{valid}}{N_{total}}
$$

Low `S_length` usually means the enzyme is cutting too aggressively or too sparsely for the target sequence.

### S_coverage

`S_coverage` is the fraction of the protein sequence covered by at least one valid peptide.

$$
S_{coverage} = \frac{R_{covered}}{L}
$$

pepVet merges overlapping valid peptide intervals before summing widths. A residue covered twice still counts once.

### S_count

`S_count` is a normalized peptide-density score.

In the current release, the expected peptide density is enzyme-aware. For digests with at least three peptide rows, pepVet uses the observed digest median length. For sparse digests, it falls back to enzyme-class expectations.

$$
N_{expected} = \frac{L}{M}
$$

$$
S_{count} = \min\left(\frac{N_{valid}}{N_{expected}}, 1\right)
$$

This is a bounded heuristic, not a mechanistic cleavage model. The change matters because a tryptic digest and a GluC digest should not be judged against the same peptide-density prior.

```{r count-demo}
ev_bsa <- evaluate_digest(bsa_path, enzyme = "trypsin", missed_cleavages = 1L)
ev_bsa$scores[, c("S_count", "median_peptide_length", "composite_score")]
```

When there are no cleavage sites for the selected enzyme, pepVet warns and sets `S_count` to `0`.

### S_hydro

`S_hydro` is the fraction of valid peptides with GRAVY values inside the active `gravy_range`.

$$
G(p) = \frac{1}{|p|}\sum_i KD_i
$$

$$
S_{hydro} = \frac{N_{window}}{N_{valid}}
$$

The default range is `-1.0` to `0.6`. The `membrane` preset widens that range because transmembrane peptides are expected to be more hydrophobic.

### S_charge

`S_charge` is the fraction of valid peptides that retain at least one non-terminal basic residue.

$$
S_{\mathrm{charge\,richness}} = \frac{N_{basic}}{N_{valid}}
$$

This score is not a claim about whether a peptide can ionize. Tryptic peptides already retain a free N-terminus and frequently a terminal Lys or Arg, so baseline ionizability is still present when `S_charge = 0`. The implementation focuses on non-terminal basic residues because those additional charge sites are what tend to support higher charge states and richer b/y ion series.

### S_unique

`S_unique` is optional. It is available only when a proteome digest is supplied.

$$
S_{unique} = \frac{N_{unique}}{N_{valid}}
$$

This score is useful for isoforms, paralogs, and family-level ambiguity. It is meaningless without a relevant proteome background, so pepVet does not invent it when `proteome` is missing.

## Weights and presets

The composite score is a weighted sum.

$$
C = \sum_i w_i S_i, \qquad \sum_i w_i = 1
$$

Default weights depend on whether `S_unique` is in play.

| Mode | Weight vector |
|---|---|---|
| Protein-only | `c(S_length = 0.200, S_coverage = 0.348, S_count = 0.226, S_hydro = 0.138, S_charge = 0.088)` |
| Proteome-aware | `c(S_length = 0.160, S_coverage = 0.279, S_count = 0.181, S_hydro = 0.110, S_charge = 0.070, S_unique = 0.200)` |

Default weights were derived using the Analytic Hierarchy Process (AHP), comparing each scoring dimension pairwise on a 1--9 Saaty scale grounded in the proteomics literature. Peptide length receives the base weight (length determines whether a peptide is detectable at all). Sequence coverage and peptide count --- both strongly correlated with digest quality --- together carry the main biological signal. Hydrophobicity and charge richness capture independent LC-MS dimensions that cannot be tuned by enzyme choice alone. The pairwise judgments are internally consistent (consistency ratio 0.03, well below the 0.10 acceptability threshold).

`pepvet_preset()` exposes curated starting points for common workflows.

```{r preset-demo}
pepvet_preset("dia")
pepvet_preset("membrane")
```

These presets modify three things together.

- `length_range`
- `gravy_range`
- `weights`

That is deliberate. Changing weights alone often hides the fact that the peptide window itself should also change.

`score_peptides()` records the resolved configuration in `preset_used`. When the active GRAVY range, peptide-length range, weights, and pI flag exactly match a shipped preset, the preset name is reported. Otherwise the call is marked `"custom"`. This is a lightweight guard against comparing scores that were generated under different workflow assumptions.

## Verdict thresholds

pepVet maps the composite score to three labels.

| Verdict | Threshold |
|---|---|
| `Good` | `>= 0.65` |
| `Moderate` | `>= 0.40` |
| `Poor` | `< 0.40` |

These thresholds are heuristic defaults. They are not calibrated to an identification probability, a false discovery rate, or a fixed instrument platform. [REF]

## Inspecting real outputs

```{r bsa-report}
digest_report(evaluate_digest(bsa_path, enzyme = "trypsin", missed_cleavages = 1L))
```

```{r h3-report}
digest_report(evaluate_digest(h3_path, enzyme = "trypsin"))
```

The BSA example usually shows a balanced score pattern. The H3.1 example often exposes the classic trypsin failure mode: many peptides, weak valid-length yield, and poor usable coverage.

## Limitations

The current model has several limits.

- It does not model post-translational modifications directly.
- It does not model chromatographic gradients, charge-state envelopes, or instrument-specific fragmentation methods.
- It uses heuristic windows and thresholds rather than empirically calibrated identification probabilities.
- It assumes the supplied proteome digest is the relevant comparison space for uniqueness.
- It assumes the active GRAVY and length windows describe a conventional C18 reversed-phase LC with ESI unless a different preset or custom configuration is supplied.
- It treats the weight vectors as expert priors, not validated coefficients.

Those limits do not make the score useless. They define what the score is for: digest planning, enzyme ranking, and transparent pre-acquisition triage.

## Scope

pepVet scores are interpretable rankings, not calibrated probabilities. The composite score has no unit. It is most honest when used to rank proteins within one enzyme-workflow configuration.

Cross-preset comparisons are not valid just because the numeric scale looks the same. A `standard` score of `0.75` and a `targeted` score of `0.60` reflect different weight vectors and can answer different questions.

```{r session-info}
sessionInfo()
```
