---
title: "Understanding the pepVet Scoring Model"
description: >
  A deep dive into the five scoring components, their mathematical definitions,
  default weight sets, verdict thresholds, and how to tune all of these for
  your specific experimental context.
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(pepVet)
bsa_path <- system.file("extdata", "P02769.fasta", package = "pepVet")
h3_path  <- system.file("extdata", "P68431.fasta", package = "pepVet")
```

## Overview

`score_peptides()` compresses a digest tibble — which can contain hundreds of
peptide rows — into a single summary row per protein. That row contains five
(or six) component scores between 0 and 1, a weighted composite score, and a
categorical verdict. Understanding what each component measures and how they
combine helps you interpret results correctly and adapt the model when the
defaults are not right for your experiment.

## The detection window

All scoring components except `S_hydro` and `S_charge` operate on **valid
peptides** only. A peptide is valid if its length is between **7 and 25
residues** (inclusive).

This window reflects the practical limits of standard reversed-phase LC-MS/MS:

- **< 7 residues**: too short to produce reliable b/y ion series; often lost
  in column void volume; retention time is irreproducible
- **> 25 residues**: precursor charge states are typically ≥ 4+ under
  standard ESI conditions; fragmentation spectra become complex and
  hard to interpret; mass accuracy requirements are strict

Peptides outside this window appear in the `digest_protein()` output (they are
real cleavage products) but they do not contribute to S_length, S_coverage, or
S_count. This is intentional — you want to know how many peptides you can
actually detect, not how many exist in principle.

```{r window-demo}
# "AAAAAAAR" = 8 residues — valid
# "AAAAAAK" = 7 residues — valid (boundary included)
# "AAAAAAR" followed immediately by another cut = 6 residues — invalid
short_seq <- "AKAKAKAKAKAKAKAKAKAKAKAKAKAKAKAKAK"  # polyAK: many 1-2 aa fragments
ev_short <- evaluate_digest(short_seq, enzyme = "trypsin")
ev_short$scores
```

## Component definitions

### S_length — peptide length distribution

**What it measures:** The fraction of cleaved peptides that fall within the
7–25 residue detection window.

$$S_{\text{length}} = \frac{|\{p \in \text{peptides} : 7 \leq |p| \leq 25\}|}{|\text{peptides}|}$$

**Interpretation:**
- 1.0 — every fragment is detectable; ideal but rare
- 0.5 — half the peptides are invisible; borderline
- < 0.3 — the enzyme over-digests or under-digests badly; expect poor
  identification rates

**What drives it down:**
- Trypsin on highly basic sequences (many consecutive K/R → many 1–3 aa
  fragments)
- Under-digesting enzymes on very large proteins (rare cuts → very long
  fragments above 25 aa)

```{r s-length}
# BSA trypsin mc=1 — high S_length because most peptides land in window
ev_bsa <- evaluate_digest(bsa_path, enzyme = "trypsin", missed_cleavages = 1L)
ev_bsa$scores[, c("S_length", "S_coverage", "S_count")]
```

### S_coverage — sequence coverage

**What it measures:** The fraction of the total protein sequence that is
covered by at least one valid peptide. Overlapping intervals are merged before
coverage is summed, so a residue covered by two overlapping peptides is counted
only once.

$$S_\text{coverage} = \frac{\text{non-overlapping residues in valid peptides}}{\text{protein length}}$$

**Interpretation:**
- 1.0 — every residue is covered by at least one detectable peptide
- 0.8+ — strong coverage; standard for well-digested, soluble proteins
- < 0.5 — large dark regions in the protein; quantification will be incomplete

**What drives it down:**
- Long stretches with no cleavage sites (e.g. transmembrane helices for
  trypsin)
- Regions generating only too-short or too-long fragments
- Protein termini that produce single sub-threshold fragments

### S_count — peptide count (soft-normalised)

**What it measures:** A normalised measure of how many valid peptides arise
from the digest, penalising both extremes — too few peptides (fragile
identification, single-peptide proteins) and too many (redundant coverage,
wasted MS time).

The current implementation uses a soft sigmoid normalisation over the
expected peptide count range for the input protein length. The score
approaches 1.0 for counts near the "ideal" density and falls toward 0 at
both extremes.

**Interpretation:**
- High scores indicate a peptide count density appropriate for reliable
  protein-level identification and quantification
- Very low scores on long proteins can indicate under-digestion
- Very low scores on short proteins (< 100 aa) often reflect the inherent
  limit of the protein size, not a problem with the enzyme

```{r s-count}
# Ubiquitin (76 aa) — very small protein, count naturally low
ubiq_path <- system.file("extdata", "P0CG48.fasta", package = "pepVet")
ev_ubiq <- evaluate_digest(ubiq_path, enzyme = "trypsin")
ev_ubiq$scores[, c("S_length", "S_coverage", "S_count", "composite_score")]
```

### S_hydro — hydrophobicity balance

**What it measures:** A score derived from the mean GRAVY (Grand Average of
Hydropathicity) of valid peptides, penalising values far from zero in either
direction.

The GRAVY score for each peptide is the mean of the per-residue Kyte-Doolittle
hydrophobicity values (stored in `aa_properties$kyte_doolittle`). A GRAVY near
zero indicates a balanced mix of hydrophilic and hydrophobic residues — optimal
for reversed-phase LC behaviour.

$$\text{GRAVY}(p) = \frac{1}{|p|} \sum_{i} \text{KD}(aa_i)$$

$$S_\text{hydro} = 1 - \frac{|\overline{\text{GRAVY}} - \text{target}|}{\text{scale}}$$

**Interpretation:**
- Score near 1.0 — mean GRAVY near 0; good reversed-phase behaviour expected
- Scores dropping below 0.5 — peptide pool is either very hydrophilic (bad
  retention, poor separation) or very hydrophobic (sticking, losses)

**What drives it down:**
- Membrane proteins with many hydrophobic transmembrane peptides push GRAVY
  very positive
- Highly acidic proteins or intrinsically disordered proteins with low
  hydrophobicity push GRAVY negative

### S_charge — ionisability

**What it measures:** The fraction of valid peptides that contain at least one
basic residue (K, R, or H). Basic residues carry positive charges under
standard ESI conditions, enabling peptide detection in positive mode.

$$S_{\text{charge}} = \frac{N_{\text{ionisable}}}{N_{\text{valid}}}$$

where $N_{\text{ionisable}}$ is the count of valid peptides containing at least one K, R, or H residue and $N_{\text{valid}}$ is the total count of valid peptides.

**Interpretation:**
- 1.0 — all detectable peptides contain at least one basic residue; ideal
- < 0.8 — a meaningful fraction of peptides are uncharged (C-terminal
  peptides, peptides from acidic regions); expect lower signal intensity
  for these peptides

**Note on terminal peptides:** The C-terminal peptide of a trypsin digest
always ends with K or R (by definition of the cut rule). However, the very
N-terminal peptide may lack K/R if the protein starts with methionine or
other non-basic residues, and this depresses S_charge slightly for trypsin
on many proteins.

### S_unique — proteome uniqueness (optional)

**What it measures:** The fraction of valid peptides that are unique to a
single protein in a supplied proteome digest. Shared peptides — those found
in more than one `protein_id` in the proteome tibble — do not contribute to
unique identification.

$$S_{\text{unique}} = \frac{N_{\text{unique}}}{N_{\text{valid}}}$$

where $N_{\text{unique}}$ is the count of valid peptides that appear in exactly one `protein_id` in the proteome and $N_{\text{valid}}$ is the total count of valid peptides.

This component is only computed when you supply a `proteome` argument to
`score_peptides()` or `evaluate_digest()`.

**Interpretation:**
- 1.0 — every peptide uniquely identifies the target protein
- < 0.7 — many peptides are shared with other proteins; protein-level
  inference will rely on a subset of peptides
- Very low values (< 0.3) — the protein is highly homologous to others in
  the proteome (paralogs, isoforms, repeated domain families); consider
  reporting at the protein group level

```{r s-unique}
# Alpha-synuclein isoforms — many shared peptides across the three isoforms
syn_path <- system.file("extdata", "P37840_isoforms.fasta", package = "pepVet")
proteome <- digest_protein(syn_path, enzyme = "trypsin")

# Evaluate isoform 1 against all three as background
ev_syn1 <- evaluate_digest(syn_path, enzyme = "trypsin",
                           proteome = proteome)
ev_syn1$scores
```

## Composite score and weights

The composite score is a weighted sum of the component scores:

$$\text{composite} = \sum_i w_i S_i, \qquad \text{where} \quad \sum_i w_i = 1$$

### Default weights — protein-only mode

When no `proteome` is supplied:

| Component | Default weight | Rationale |
|---|---|---|
| `S_length` | 0.25 | Primary filter — useless peptides cost everything downstream |
| `S_coverage` | 0.25 | Sequence coverage is the top-line quality metric in proteomics |
| `S_count` | 0.20 | More peptides = more identification evidence |
| `S_hydro` | 0.15 | Important but often acceptable in a moderate range |
| `S_charge` | 0.15 | High priority but most tryptic digests already score well here |

### Default weights — proteome-aware mode

When a `proteome` is supplied, `S_unique` is added and the weights are
rebalanced to give uniqueness a meaningful share without completely dominating:

| Component | Default weight |
|---|---|
| `S_length` | 0.20 |
| `S_coverage` | 0.20 |
| `S_count` | 0.15 |
| `S_hydro` | 0.15 |
| `S_charge` | 0.10 |
| `S_unique` | 0.20 |

### Customising weights

You can supply your own weight vector via the `weights` argument. The vector
must be named with the component names you want to weight, and pepVet will
normalise the weights to sum to 1 automatically.

```{r custom-weights}
# Downweight hydrophobicity (your column does well with everything)
# and upweight coverage (your protein has known dark regions)
custom_weights <- c(
  S_length   = 0.25,
  S_coverage = 0.35,
  S_count    = 0.20,
  S_hydro    = 0.05,
  S_charge   = 0.15
)

digest_result <- digest_protein(bsa_path, enzyme = "trypsin",
                                missed_cleavages = 1L)
score_peptides(digest_result, weights = custom_weights)
```

**When to customise weights:**

- **Targeted quantification (SRM/PRM)**: `S_coverage` matters less (you
  already know which peptides to target); `S_charge` and `S_hydro` matter
  more.
- **Isoform discrimination**: give `S_unique` a higher weight (0.30–0.40)
  when allele or isoform resolution is the primary goal.
- **DIA-based label-free quantification**: `S_count` can be weighted more
  heavily because more peptides per protein improves the library matching
  confidence.
- **Membrane proteomics**: down-weight `S_hydro` (you expect hydrophobic
  peptides) and focus on `S_length` and `S_coverage`.

## Verdict thresholds

Verdicts map the continuous composite score to three categories:

| Verdict | Threshold | Practical meaning |
|---|---|---|
| **Good** | composite ≥ 0.70 | Expect reliable identification and quantification with standard methods |
| **Moderate** | composite ≥ 0.40 | Expect partial coverage; identification is possible but not robust |
| **Poor** | composite < 0.40 | Unreliable; consider a different enzyme, sample prep strategy, or fractional digestion |

These thresholds are heuristic defaults tuned against manually reviewed
proteomics datasets. They are not calibrated against a specific FDR or
identification rate model. See `plan/research.md` for the current empirical
validation status and the roadmap for calibrating against PeptideAtlas data
(planned for Commit 16).

**Important caveat:** A Good verdict does not guarantee identification. It
indicates that the peptide set properties are consistent with what
well-characterised Good proteins look like. Sample quality, instrument
settings, search parameters, and protein abundance all affect the final result
independently of the digest quality.

## Inspecting component scores interactively

`digest_report()` renders the component scores as a horizontal bar chart,
making it easy to see immediately which component is dragging the composite
down.

```{r report-bsa}
ev_bsa <- evaluate_digest(bsa_path, enzyme = "trypsin", missed_cleavages = 1L)
digest_report(ev_bsa)
```

```{r report-h3}
ev_h3 <- evaluate_digest(h3_path, enzyme = "trypsin")
digest_report(ev_h3)
```

On H3.1, `S_length` and `S_coverage` are both very low while `S_hydro` is
near-perfect. This immediately tells you the problem: trypsin over-digests the
protein, producing many correctly hydrophilic fragments that are just too short
to detect. The enzyme — not the hydrophobicity, not the charge — is the
problem, and switching to LysC directly addresses it.

## Edge cases and known limitations

**Very short proteins (< 50 aa):** S_count will be low by definition because
there are simply not enough residues to generate many valid peptides. This
is not a digest quality problem. Ubiquitin (76 aa) typically scores Moderate
with any enzyme.

**Repeated sequence proteins:** Proteins with large internal repeats (ankyrin
domains, WD40 repeats, collagen repeats) produce many highly similar peptides.
S_unique will be low when scored against a proteome containing the same repeat
family. This is a genuine proteomics challenge, not a scoring artefact.

**PTM-blind scoring:** The current model scores the unmodified primary sequence.
Phosphorylation, acetylation, methylation or other PTMs that alter cleavage
efficiency (e.g. missed cleavages near phosphoserine) are not modelled. This is
documented in `plan/research.md` and targeted in Commit 17.

**S_count normalisation is protein-length relative:** The soft-sigmoid
normalisation in S_count uses the input protein length to set the expected
peptide density. Very long proteins (like titin, 34,350 aa) and very short
proteins (like ubiquitin, 76 aa) are both normalised relative to their own
length, so cross-protein comparisons of S_count alone are not meaningful in
absolute terms — use the composite score for cross-protein comparison.

```{r session-info}
sessionInfo()
```
