---
title: "Score Diagnostics for pepVet Scoring Models"
description: >
  Use score_diagnostics() to assess multicollinearity, dimensionality,
  and component importance in pepVet scoring models. A guide to VIF, PCA,
  and ablation analysis with worked examples.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Score Diagnostics for pepVet Scoring Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(pepVet)
small_fasta <- system.file(
  "extdata", "small_proteome_50_proteins.fasta",
  package = "pepVet"
)
```

pepVet's scoring model combines five (or six) component scores into a weighted composite. The component scores are designed to measure orthogonal digest-quality dimensions, but correlations among them are not guaranteed by the formula. `score_diagnostics()` quantifies three properties of the scoring model: multicollinearity, effective dimensionality, and the marginal contribution of each component.

This article covers the three analyses, how to interpret them, and what to do if the diagnostics reveal issues.

## Setup

Evaluate the 50-protein fixture with trypsin. This gives a batch result with five component scores per protein.

```{r batch}
batch <- batch_evaluate(small_fasta, enzyme = "trypsin")
dim(batch)
```

## Running diagnostics

A single call to `score_diagnostics()` runs all three analyses:

```{r diag}
d <- score_diagnostics(batch)
```

The result is a list with six elements. The three analysis results are in `$vif`, `$pca`, and `$ablation`.

```{r structure}
names(d)
```

## VIF: multicollinearity

Variance Inflation Factor measures how much the variance of each component score is inflated by linear dependence on the other components. For each component, we fit a linear model predicting it from the remaining components and compute VIF = 1 / (1 - R^2).

VIF < 5 is acceptable. VIF between 5 and 10 is concerning. VIF > 10 indicates problematic collinearity. These thresholds are standard statistical practice [REF: O'Brien 2007, Quality & Quantity 41:673].

```{r vif}
d$vif
```

Components with VIF > 5 are flagged. In this fixture, S_length has VIF = 5.78, slightly above the concerning threshold. This makes sense: peptide length correlates with coverage (longer proteins tend to have more valid peptides covering more of the sequence). The correlation is moderate and does not require corrective action.

## PCA: dimensionality

Principal Component Analysis reveals how many effective dimensions the component scores occupy. If PC1 + PC2 explain more than 80% of variance, the scoring model has low effective dimensionality. Most of the variation across proteins is captured by two composite patterns.

```{r pca}
round(d$pca$var_explained, 3)
```

In this fixture, PC1 + PC2 explain `r round(sum(d$pca$var_explained[1:2]) * 100)`% of variance. The scree plot shows one dominant PC followed by a gradual decline. This is the expected pattern for a well-designed multi-criteria score: components share some covariance (because they all measure aspects of digest quality), but each retains unique information.

The loadings matrix shows which original components contribute to each PC. PC1 is a weighted average of all five components (all loadings positive, ranging from 0.34 to 0.50). PC2 is driven primarily by hydrophobicity (loading -0.86), capturing the dimension that is most orthogonal to the others.

```{r loadings, echo = FALSE}
knitr::kable(round(d$pca$loadings, 3),
  caption = "PCA loadings (rotation matrix)")
```

## Ablation: component importance

Ablation measures what happens to the composite score when each component is set to zero. This answers: "if this component contributed nothing, how much would the composite drop, and how many proteins would change verdict?"

```{r ablation}
d$ablation
```

Key patterns to look for:

- **Mean drop**: components with large weights should produce proportionally large drops. S_coverage (weight 0.348) produces the largest mean drop (0.240). S_charge (weight 0.088) produces the smallest (0.062). This confirms the weights are consistent with the data.

- **SD drop**: components with high variability across proteins produce larger SD. S_coverage varies the most (SD 0.061). This is expected: coverage depends on how well the enzyme cuts each specific protein.

- **Verdict flips**: the number of proteins whose verdict changes when that component is removed. S_coverage removal flips 45 of 50 verdicts. S_charge removal flips only 14. This tells you which components are most critical for the final verdict classification.

## The diagnostics plot

`plot_score_diagnostics()` produces a three-panel figure combining all three analyses:

```{r plot, fig.width = 9.6, fig.height = 3, dpi = 150}
if (requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("patchwork", quietly = TRUE)) {
  p <- plot_score_diagnostics(d)
  print(p)
}
```

**Panel A (VIF):** Bars colored by severity. Dashed lines at 5 and 10 mark the concerning and problematic thresholds. Components in orange or red should be investigated. Consider whether they are measuring a distinct dimension or could be merged with a correlated component.

**Panel B (PCA):** Bar heights show the variance explained by each PC. The cumulative line (with points) shows total variance explained up to each PC. The dashed 80% reference helps assess dimensionality. The subtitle reports PC1+PC2 variance as a single summary number.

**Panel C (Ablation):** Bars show the mean composite drop when each component is set to zero. Error bars show +/- 1 SD. The annotation shows the exact drop value and the number of verdict flips (out of total proteins). Components with small drops and few flips are candidates for down-weighting. Components with large drops and many flips are critical to the verdict.

## Interpreting the results

### What good diagnostics look like

- All VIF values below 5.
- PC1 + PC2 above 80%.
- Ablation mean drops proportional to component weights.
- No component has zero verdict flips (every component contributes something).

### What to do if VIF is high

If a component has VIF > 10, it shares most of its variance with other components. Options:

1. Remove the redundant component and redistribute its weight to the others.
2. Keep the component and document the collinearity as a limitation.

In practice, modest VIF values (5-10) are expected because all components measure aspects of digest quality. The AHP framework accounts for this by design. The pairwise comparisons explicitly consider that dimensions overlap.

### When to run diagnostics

Run `score_diagnostics()` whenever you:

- Change the default weights or add a new component.
- Want to check that the scoring model behaves as expected on a new proteome.
- Need evidence for a manuscript that the component scores are not redundant.
- Suspect that two components may be measuring the same thing (for example, if S_count and S_coverage behave similarly on a particular enzyme).

## Session info

```{r session-info}
sessionInfo()
```
