---
title: "Visualising Digest Quality"
description: >
  Use pepVet's plotting functions to inspect single-protein digests,
  compare enzymes, explore physicochemical distributions, and assess
  proteome-scale performance.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %VignetteIndexEntry{Visualising Digest Quality}
  %VignetteEngine{knitr::rmarkdown}
  %VignetteEncoding{UTF-8}
---

## Overview

pepVet provides 11 visualisation functions organised by analytical scope.
Every function returns a `ggplot` or `patchwork` object that can be
further customised, combined, or saved with `pepvet_save_figure()`.

```{r setup, eval = requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("patchwork", quietly = TRUE)}
library(pepVet)
library(patchwork)
```

---

## Single-protein diagnostics

### Four-panel digest profile

`plot_digest_profile()` is the flagship function. It assembles length
distribution, GRAVY hydrophobicity, sequence coverage, and component
scores into one publication-ready figure.

```{r digest-profile, fig.width=12, fig.height=10}
bsa_path <- system.file("extdata", "P02769.fasta", package = "pepVet")
bsa_trypsin <- evaluate_digest(bsa_path, enzyme = "trypsin")
plot_digest_profile(bsa_trypsin)
```

### Coverage map

`plot_coverage_map()` shows how peptides map onto the protein sequence,
with separate lanes for each missed-cleavage level.

```{r coverage-map, fig.width=10, fig.height=6}
plot_coverage_map(bsa_trypsin, color_by = "length_class")
```

### Cleavage map

`plot_cleavage_map()` marks every cleavage site and highlights
efficiency when annotation data is available.

```{r cleavage-map, fig.width=10, fig.height=4}
cs <- annotate_cleavage_sites(bsa_path, enzyme = "trypsin")
plot_cleavage_map(bsa_trypsin, cleavage_sites = cs)
```

### Peptide overlap map

`plot_peptide_overlap_map()` colours each residue by how many
detected peptides cover it. Useful for assessing redundancy.

```{r overlap-map, fig.width=12, fig.height=8}
plot_peptide_overlap_map(bsa_trypsin)
```

---

## Enzyme & protein comparison

### Enzyme comparison

`plot_enzyme_comparison()` compares multiple enzymes on the same
protein using component-score bars and a composite lollipop.

```{r enzyme-comp, fig.width=10, fig.height=7}
comp <- compare_digests(bsa_path,
  enzymes = c("trypsin", "lysc", "chymotrypsin-high"))
plot_enzyme_comparison(comp)
```

---

## Physicochemical distributions

### Peptide length distribution

```{r length-dist, fig.width=8, fig.height=6}
plot_length_distribution(bsa_trypsin)
```

### GRAVY landscape

The 2D scatter of length vs. GRAVY reveals physicochemical outliers.

```{r gravy-landscape, fig.width=10, fig.height=8}
plot_gravy_landscape(bsa_trypsin)
```

### pI distribution

Preview fractionation outcomes with SCX fraction bins.

```{r pi-dist, fig.width=8, fig.height=6}
pi_data <- score_peptides(bsa_trypsin$peptides,
  enzyme = "trypsin", include_pI = TRUE)
plot_pI_distribution(pi_data)
```

### m/z distribution

Check whether your peptides fall within the instrument scan range.

```{r mz-dist, fig.width=8, fig.height=6}
plot_mz_distribution(bsa_trypsin)
```

### Missed-cleavage impact

See how allowing more missed cleavages changes each score component.

```{r mc-impact, fig.width=8, fig.height=6}
mc0 <- evaluate_digest(bsa_path, enzyme = "trypsin", missed_cleavages = 0)
mc1 <- evaluate_digest(bsa_path, enzyme = "trypsin", missed_cleavages = 1)
mc2 <- evaluate_digest(bsa_path, enzyme = "trypsin", missed_cleavages = 2)
plot_missed_cleavage_impact(list("MC=0" = mc0, "MC=1" = mc1, "MC=2" = mc2))
```

---

## Proteome-scale views

Load the 50-protein test proteome and evaluate it with 10 diverse enzymes.
`batch_evaluate()` uses parallel workers when available (set `cores` to your
CPU count for faster execution).

```{r batch, message=FALSE, eval = requireNamespace("Biostrings", quietly = TRUE)}
library(Biostrings)
fasta_path <- system.file("extdata", "small_proteome_50_proteins.fasta",
                           package = "pepVet")
proteome <- readAAStringSet(fasta_path)
cat(sprintf("Loaded %d proteins\n", length(proteome)))

enzymes <- c("trypsin", "lysc", "chymotrypsin-high",
             "asp-n endopeptidase", "glutamyl endopeptidase",
             "arg-c proteinase", "thermolysin", "pepsin",
             "staphylococcal peptidase i", "proteinase k")
cat(sprintf("Evaluating %d enzymes ...\n", length(enzymes)))

# Evaluate one enzyme for the proteome overview
batch_trypsin <- batch_evaluate(proteome, enzyme = "trypsin",
                                cores = 1L)
```

### Proteome overview

Three-panel portrait for trypsin: score distribution, component profile,
and difficulty flags.

```{r proteome-overview, fig.width=12, fig.height=10, eval = exists("batch_trypsin")}
plot_proteome_overview(batch_trypsin)
```

### Batch enzyme comparison

Compare all 10 enzymes across the 50-protein proteome.
For production use, increase `cores` to match your CPU count.

```{r batch-comp, fig.width=16, fig.height=12, eval = exists("proteome")}
batch_comp <- batch_compare_enzymes(proteome, enzymes = enzymes, cores = 1L)
plot_batch_comparison(batch_comp,
  title = sprintf("50-Protein Proteome  \u00b7  %d Enzymes", length(enzymes)))
```

---

## Customising and exporting plots

All pepVet plots return standard ggplot2 objects. Add theme layers,
modify scales, or compose with patchwork.

```{r custom, fig.width=10, fig.height=7}
plot_length_distribution(bsa_trypsin) +
  ggplot2::labs(subtitle = "My custom subtitle") +
  pepvet_theme_presentation()
```

Use `pepvet_save_figure()` for publication-ready output:

```{r save, eval=FALSE}
p <- plot_digest_profile(bsa_trypsin)
pepvet_save_figure(p, "bsa_digest.png")
```

Override colors and thresholds globally with `pepvet_plot_config()`:

```{r config, eval=FALSE}
pepvet_plot_config(
  palette = list(brand = "#004488"),
  params  = list(verdict_good = 0.70)
)
pepvet_plot_config_reset()
```
