---
title: "Weight Sensitivity Analysis"
description: >
  Measure how pepVet verdicts and enzyme rankings change under Dirichlet-based Monte Carlo perturbations of the scoring weights.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Weight Sensitivity Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(pepVet)
```

## Analysis scope

pepVet's scoring weights (S_length = 0.200, S_coverage = 0.348, S_count = 0.226, S_hydro = 0.138, S_charge = 0.088) determine how each component contributes to the composite. `sensitivity_analysis()` perturbs these weights and records how often the verdict or enzyme ranking changes. It asks whether the result is stable within the specified perturbation model.

A verdict that is unchanged in more than 95% of perturbations is stable to the tested weight changes. This does not show that the weights or verdict are experimentally correct. A less stable result indicates that the score lies close enough to a decision boundary for modest changes in weighting to alter the label.

## Single-protein examples

The BSA and Histone H3.1 examples contrast a stable result with one that changes near a verdict boundary under the same perturbation model.

### A stable case: BSA with trypsin

```{r bsa}
bsa_path <- system.file("extdata", "P02769.fasta", package = "pepVet")
res <- evaluate_digest(bsa_path, enzyme = "trypsin", missed_cleavages = 1L)
sens <- sensitivity_analysis(res, n_iter = 5000L)
```

The default composite for BSA trypsin is 0.885, above the 0.65 Good threshold. The simulation records the verdict in each weight draw:

```{r bsa-verdict}
sens$summary$verdict_pct
```

All 5,000 draws return `Good`. The central 95% simulation interval for the composite is `r round(sens$summary$composite_ci[1], 3)` to `r round(sens$summary$composite_ci[2], 3)`. This result establishes stability only for the configured Dirichlet distribution and random seed.

### A borderline case: Histone H3.1 with trypsin

```{r histone}
h3_path <- system.file("extdata", "P68431.fasta", package = "pepVet")
res_h3 <- evaluate_digest(h3_path, enzyme = "trypsin", missed_cleavages = 1L)
sens_h3 <- sensitivity_analysis(res_h3, n_iter = 5000L)
```

Histone H3.1 is lysine- and arginine-rich. Strict tryptic digestion produces many short products, and the default composite sits near the 0.65 boundary. The sensitivity analysis measures how often the label changes under the specified weight perturbations:

```{r histone-verdict}
sens_h3$summary$verdict_pct
```

The verdict changes between `Good` and `Moderate` across the sampled weights. The result is sensitive to weighting near the 0.65 boundary. Compare other enzymes if that sensitivity affects the decision, but evaluate them under the same scoring configuration.

### Visualising single-protein verdict distributions

```{r plot, fig.width=7, fig.height=4}
p <- plot_weight_sensitivity(sens_h3)
p
```

The density spans both sides of the 0.65 boundary. In the BSA simulation, all sampled composites remain above that boundary:

```{r plot-bsa, fig.width=7, fig.height=3}
p_bsa <- plot_weight_sensitivity(sens)
p_bsa
```

## Proteome-scale analysis: 50 proteins across 10 enzymes

The batch example applies the same perturbation to 50 proteins and 10 enzymes.

```{r batch}
small_path <- system.file("extdata", "small_proteome_50_proteins.fasta",
  package = "pepVet")
enzymes <- c("trypsin", "lysc", "chymotrypsin-high", "asp-n endopeptidase",
  "glutamyl endopeptidase", "arg-c proteinase", "thermolysin",
  "pepsin", "staphylococcal peptidase i", "proteinase k")
batch_comp <- batch_compare_enzymes(small_path, enzymes = enzymes)
sens <- sensitivity_analysis(batch_comp, n_iter = 2000L)
```

Overall instability across the entire proteome:

```{r total-inst}
sens$summary$total_instability
```

This means `r round(sens$summary$total_instability * 100, 1)`% of all protein-enzyme-iteration triples produced a different verdict than the default weights would give.

### Pairs with fewer than 5% verdict changes

```{r stable-fraction}
mean(sens$per_protein$verdict_instability < 0.05)
```

`r round(mean(sens$per_protein$verdict_instability < 0.05) * 100)`% of protein-enzyme pairs have a verdict that changes in fewer than 5% of iterations. These verdicts are stable under the tested perturbations, which is narrower than evidence of experimental validity.

### Scores near verdict boundaries

The rows with the largest instability have mean composites near the 0.65 or 0.40 thresholds in this simulation.

```{r worst-proteins}
worst <- head(sens$per_protein[order(sens$per_protein$verdict_instability,
  decreasing = TRUE), ], 10)
worst[, c("protein_id", "verdict_instability", "composite_mean",
  "composite_lo", "composite_hi")]
```

The displayed rows have simulation intervals that cross a verdict boundary, matching the pattern in the H3.1 example.

### Instability by enzyme

Mean verdict instability is the fraction of iterations in which the verdict differs from the default-weight verdict. A value of 0.05 means 5% of iterations changed the label.

```{r by-enzyme}
sort(tapply(sens$per_protein$verdict_instability,
  sens$per_protein$enzyme, mean))
```

For trypsin, a mean near 0.08 means that 8% of the sampled iterations changed the verdict, averaged across proteins. The displayed means describe this fixture and perturbation model. They do not isolate why enzymes differ.

### Visualising the batch results

```{r batch-plot, fig.width=10, fig.height=6}
p_batch <- plot_weight_sensitivity(sens)
p_batch
```

Each panel shows per-protein verdict instability for one enzyme. Values near zero changed label rarely under the sampled weights. Larger values identify protein-enzyme pairs that depend more on the weight vector.

## Interpretation

### What the analysis does and does not tell you

It tells you whether the verdict or ranking is stable to the tested weight values. At single-protein scale, it distinguishes a result that is unchanged in this simulation (BSA with trypsin, 100% Good) from one that changes frequently (H3.1 with trypsin, about 15% Good and 85% Moderate). At proteome scale, it reports the fraction in each category and which enzymes have fewer boundary-sensitive cases in the supplied data.

It does not tell you whether the weights are correct in an absolute sense. That would require empirical calibration against observed peptide detection rates. It does not tell you whether the verdict thresholds (0.65, 0.40) are optimal. Those are held fixed during the analysis.

The practical use is to identify results that depend on the chosen weights. High stability means the label changed rarely under this perturbation scheme. High instability means that the component weighting deserves closer inspection. Neither result establishes experimental success or failure.

### Technical notes

The perturbation uses a Dirichlet distribution with concentration parameters alpha_i = nu * w0_i (nu = 63). This gives a standard deviation of approximately 0.05 for a weight of 0.2. Dirichlet draws are non-negative and sum to 1, so they preserve the weight simplex without clamping and renormalizing independent draws. The method follows a simplified stochastic multicriteria acceptability approach (Lahdelma et al. 1998, Tervonen and Figueira 2008).

For batch inputs, the analysis processes proteins in chunks of 10,000 to bound the working set.

Proteins with no cleavage sites for a given enzyme receive composite = 0 and verdict = "Poor" as a hard override. This ensures undigestible proteins remain negative controls under all weight perturbations.

## References

Butler J, Jia J, Dyer J. Simulation techniques for the sensitivity analysis of multi-criteria decision models. *European Journal of Operational Research*. 1997;103:531-546.

Lahdelma R, Hokkanen J, Salminen P. SMAA: Stochastic multiobjective acceptability analysis. *European Journal of Operational Research*. 1998;106(1):137-143.

Mazurek J, Strzalka D. On the Monte Carlo weights in multiple criteria decision analysis. *PLOS ONE*. 2022;17(10):e0268950.

Tervonen T, Figueira JR. A survey on stochastic multicriteria acceptability analysis methods. *Journal of Multi-Criteria Decision Analysis*. 2008;15(1-2):1-14.

## Session info

```{r session-info}
sessionInfo()
```
