---
title: "Workflow Presets in pepVet"
description: >
  A practical guide to pepVet presets for standard DDA, DIA, targeted assays,
  membrane proteins, degraded material, and fractionated workflows.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Workflow Presets in pepVet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(pepVet)
bsa_path <- system.file("extdata", "P02769.fasta", package = "pepVet")
syn_path <- system.file("extdata", "P37840_isoforms.fasta", package = "pepVet")
```

## Why presets exist

Most digest-scoring tools expose a fixed set of assumptions. Real experiments do not. A targeted assay, a membrane-protein workflow, and an FFPE rescue experiment do not value the same peptide properties to the same degree.

`pepvet_preset()` packages three related decisions into one object.

- the valid peptide-length window
- the acceptable GRAVY window
- the component weight vector

That keeps the scoring logic explicit and easy to modify.

`score_peptides()` records the resolved workflow label in `preset_used`. When a call exactly matches one of pepVet's shipped presets, that preset name is reported. Otherwise the call is marked `"custom"`.

## Available presets

```{r list-standard}
pepvet_preset("standard")
```

| Preset | Intended use | Key change | Literature basis |
|---|---|---|---|---|
| `standard` | Routine DDA | Default AHP weights | Tabb 2008 |
| `dia` | DIA and SWATH | `[7,30]` aa, GRAVY `[-1,0.8]`, high coverage | Ludwig et al. 2018 |
| `targeted` | SRM, PRM, MRM | `[8,20]` aa, GRAVY `[-0.8,0.4]`, S_unique 30% | Lange et al. 2008, Picotti & Aebersold 2012 |
| `membrane` | Hydrophobic proteins | length `[7,30]`, GRAVY `[-1.0,2.0]`, S_hydro 5% | Vit & Petrak 2017 |
| `ffpe_degraded` | Degraded material | `[6,30]` aa, high S_count weight | Coscia et al. 2020, Buczak et al. 2023 |
| `fractionated` | Offline fractionation | Standard + pI annotation | - |

## Standard workflow

The `standard` preset is a clear baseline for routine peptide-centric proteomics.

```{r standard-example}
standard <- pepvet_preset("standard")

do.call(
  evaluate_digest,
  c(list(sequence = bsa_path, enzyme = "trypsin", missed_cleavages = 1L), standard)
)$scores
```

## DIA workflow

DIA benefits from broad peptide support across the protein because fragmentation is decoupled from precursor selection. The `dia` preset shifts weight toward coverage, widens the valid length window to 30 aa, and extends the GRAVY window to \[-1.0, 0.8\] since DIA tolerates more hydrophobic peptides (Ludwig et al. 2018, *Cell Systems*).

```{r dia-example}
syn_proteome <- digest_protein(syn_path, enzyme = "trypsin")
dia <- pepvet_preset("dia")

do.call(
  evaluate_digest,
  c(list(sequence = syn_path, enzyme = "trypsin", proteome = syn_proteome), dia)
)$scores
```

## Targeted workflow

Targeted assays care about a smaller set of peptides, but those peptides need to be stable and specific. The `targeted` preset pushes more weight onto `S_unique` and expects a relevant comparison proteome.

```{r targeted-example}
proteome <- digest_protein(syn_path, enzyme = "trypsin")
targeted <- pepvet_preset("targeted")

do.call(
  evaluate_digest,
  c(list(sequence = syn_path, enzyme = "trypsin", proteome = proteome), targeted)
)$scores
```

Without a proteome digest, a non-zero uniqueness weight is not valid. pepVet raises an error instead of pretending the score can be computed.

## Membrane workflow

Membrane proteins are a common reason to reject default hydrophobicity assumptions. The `membrane` preset extends the GRAVY window to \[-1.0, 2.0\] and the length window to \[7, 30\] because transmembrane helices contain long hydrophobic stretches with GRAVY values far exceeding soluble proteins (Vit & Petrak 2017, *J Proteomics*). The S_hydro weight drops to 5% to avoid penalising the very hydrophobicity that defines this class.

```{r membrane-example}
pepvet_preset("membrane")
```

This does not guarantee a membrane digest is good. It only stops the scoring model from treating all hydrophobic peptides as implausible by default.

## Degraded samples

The `ffpe_degraded` preset lowers the minimum length to 6 aa because FFPE-derived peptides are shorter due to formalin-induced cross-linking and degradation (Coscia et al. 2020, *Nat Commun*; Buczak et al. 2023, *Mol Cell Proteomics*). The elevated S_count weight prioritises having more quantifiable peptides to compensate for reduced overall signal.

```{r ffpe-example}
pepvet_preset("ffpe_degraded")
```

Use this preset only when the sample context justifies it. A wider window changes the meaning of `S_length`, `S_coverage`, and `S_count`.

## Fractionated workflows

The `fractionated` preset keeps a conventional scoring profile but is useful when planning offline fractionation. The digest still needs strong peptides, but you may tolerate a broader peptide inventory because the acquisition strategy is designed to spread complexity over several runs.

## Editing a preset

Presets are plain R lists. The easiest way to customize them is to start with one and change the fields you need.

```{r modify-preset}
custom_fractionated <- pepvet_preset("fractionated")
custom_fractionated$weights["S_coverage"] <- 0.30
custom_fractionated$weights["S_hydro"] <- 0.10
custom_fractionated$weights <-
  custom_fractionated$weights / sum(custom_fractionated$weights)

do.call(
  evaluate_digest,
  c(list(sequence = bsa_path, enzyme = "chymotrypsin-high"), custom_fractionated)
)$scores
```

This is deliberate API design. Presets are data, not hidden state. You can inspect them, store them, version them, and adapt them to local methods without rewriting package internals.

## Real workflow examples

The presets are most useful in three situations.

- You need a defensible starting point for a new assay.
- You want comparisons to stay consistent across a panel of proteins.
- You need to explain why a score changed when the workflow changed.

That third case matters in practice. If a membrane target improves only after widening the GRAVY window, the result is interpretable. The score changed because the workflow assumptions changed, not because the package silently drifted.

## Cross-preset interpretation

Do not compare composite scores across different presets as if they were on one universal scale. A `standard` score and a `targeted` score can differ because the weight vector, peptide-length window, and GRAVY window changed, not because the protein became intrinsically better or worse.

Use `preset_used` as a guardrail when you save or compare results. Scores are directly comparable only when the resolved scoring configuration matches.

## Summary

Use presets as starting points, not as immutable truth. They are there to make workflow assumptions explicit, reproducible, and easy to modify.

```{r session-info}
sessionInfo()
```
