---
title: "Workflow Presets in pepVet"
description: >
  A practical guide to pepVet presets for standard DDA, DIA, targeted assays, membrane proteins, degraded material, and fractionated workflows.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Workflow Presets in pepVet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(pepVet)
bsa_path <- system.file("extdata", "P02769.fasta", package = "pepVet")
syn_path <- system.file("extdata", "P37840_isoforms.fasta", package = "pepVet")
```

## Why presets exist

Most digest-scoring tools expose a fixed set of assumptions. Real experiments do not. A targeted assay, a membrane-protein workflow, and an FFPE rescue experiment do not value the same peptide properties to the same degree.

`pepvet_preset()` packages three related decisions into one object.

- the valid peptide-length window
- the acceptable GRAVY window
- the component weight vector

The returned object keeps the scoring settings explicit and editable.

`score_peptides()` records the resolved workflow label in `preset_used`. When a call exactly matches one of pepVet's shipped presets, the function reports that preset name. Otherwise it reports `"custom"`.

## Available presets

```{r list-standard}
pepvet_preset("standard")
```

| Preset | Intended use | Key change |
|---|---|---|
| `standard` | Routine DDA | Default AHP weights |
| `dia` | DIA and SWATH | `[7,30]` aa, GRAVY `[-1,0.8]`, high coverage weight |
| `targeted` | SRM, PRM, MRM | `[8,20]` aa, GRAVY `[-0.8,0.4]`, `S_unique` 30% |
| `membrane` | Hydrophobic proteins | Length `[7,30]`, GRAVY `[-1.0,2.0]`, `S_hydro` 5% |
| `ffpe_degraded` | Degraded material | `[6,30]` aa, higher `S_count` weight |
| `fractionated` | Offline fractionation | Standard settings with pI annotation |

## Standard workflow

The `standard` preset contains the package defaults for routine peptide-centric proteomics.

```{r standard-example}
standard <- pepvet_preset("standard")

do.call(
  evaluate_digest,
  c(list(sequence = bsa_path, enzyme = "trypsin", missed_cleavages = 1L), standard)
)$scores
```

## DIA workflow

The `dia` preset assigns more weight to coverage, widens the valid length window to 30 aa, and extends the GRAVY window to \[-1.0, 0.8\]. These values encode a broader inclusion prior for DIA workflows. They are not calibrated detection boundaries.

```{r dia-example}
syn_proteome <- digest_protein(syn_path, enzyme = "trypsin")
dia <- pepvet_preset("dia")

do.call(
  evaluate_digest,
  c(list(sequence = syn_path, enzyme = "trypsin", proteome = syn_proteome), dia)
)$scores
```

## Targeted workflow

The `targeted` preset assigns more weight to `S_unique` and requires a relevant comparison proteome. It uses a narrower length and GRAVY window than the standard preset.

```{r targeted-example}
proteome <- digest_protein(syn_path, enzyme = "trypsin")
targeted <- pepvet_preset("targeted")

do.call(
  evaluate_digest,
  c(list(sequence = syn_path, enzyme = "trypsin", proteome = proteome), targeted)
)$scores
```

Without a proteome digest, pepVet cannot calculate a non-zero uniqueness component and raises an error.

## Membrane workflow

The `membrane` preset extends the GRAVY window to \[-1.0, 2.0\], extends the length window to \[7, 30\], and lowers the `S_hydro` weight to 5%. These settings reduce the score penalty for hydrophobic peptides. They remain workflow priors and do not predict peptide observation.

```{r membrane-example}
pepvet_preset("membrane")
```

The preset changes how the score treats hydrophobic peptides. It does not predict whether the digest will succeed experimentally.

## Degraded samples

The `ffpe_degraded` preset lowers the minimum length to 6 aa and increases the `S_count` weight. This configuration includes more short products and gives peptide count more influence on the composite score. It does not model formalin chemistry or degradation directly.

```{r ffpe-example}
pepvet_preset("ffpe_degraded")
```

Use this preset only when the sample context justifies it. A wider window changes the meaning of `S_length`, `S_coverage`, and `S_count`.

## Fractionated workflows

The `fractionated` preset keeps the standard score settings and requests pI annotation for planning offline fractionation.

## Editing a preset

Presets are plain R lists. Copy one and change the fields required by the workflow.

```{r modify-preset}
custom_fractionated <- pepvet_preset("fractionated")
custom_fractionated$weights["S_coverage"] <- 0.30
custom_fractionated$weights["S_hydro"] <- 0.10
custom_fractionated$weights <-
  custom_fractionated$weights / sum(custom_fractionated$weights)

do.call(
  evaluate_digest,
  c(list(sequence = bsa_path, enzyme = "chymotrypsin-high"), custom_fractionated)
)$scores
```

Presets are data, not hidden state. You can inspect, store, version, and modify them without changing package internals.

## When presets help

Presets provide a recorded starting configuration when:

- You need a starting point for a new assay.
- You want comparisons to stay consistent across a panel of proteins.
- You need to explain why a score changed when the workflow changed.

If a membrane target receives a higher score only after widening the GRAVY window, the recorded configuration shows which assumption changed.

## Cross-preset interpretation

Do not compare composite scores across different presets as if they were on one universal scale. A `standard` score and a `targeted` score can differ because the weight vector, peptide-length window, and GRAVY window changed, not because the protein became intrinsically better or worse.

Use `preset_used` when saving or comparing results. Compare scores only when the resolved scoring configuration matches.

Presets are starting points, not experimentally fitted parameter sets. Modify them when the workflow requires different assumptions.

```{r session-info}
sessionInfo()
```
