---
title: "Workflow Presets in pepVet"
description: >
  A practical guide to pepVet presets for standard DDA, DIA, targeted assays,
  membrane proteins, degraded material, and fractionated workflows.
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(pepVet)
bsa_path <- system.file("extdata", "P02769.fasta", package = "pepVet")
syn_path <- system.file("extdata", "P37840_isoforms.fasta", package = "pepVet")
```

## Why presets exist

Most digest-scoring tools expose a fixed set of assumptions. Real experiments do not. A targeted assay, a membrane-protein workflow, and an FFPE rescue experiment do not value the same peptide properties to the same degree.

`pepvet_preset()` packages three related decisions into one object.

- the valid peptide-length window
- the acceptable GRAVY window
- the component weight vector

That keeps the scoring logic explicit and easy to modify.

## Available presets

```{r list-standard}
pepvet_preset("standard")
```

| Preset | Intended use | Main change |
|---|---|---|
| `standard` | Routine DDA | Package defaults |
| `dia` | DIA and SWATH | Higher coverage weight |
| `targeted` | SRM, PRM, MRM | Higher uniqueness weight |
| `membrane` | Hydrophobic proteins | Wider GRAVY window |
| `ffpe_degraded` | Degraded or fragmented material | Wider length window |
| `fractionated` | Offline fractionation planning | Standard weights with fractionation-oriented interpretation |

## Standard workflow

The `standard` preset is a clear baseline for routine peptide-centric proteomics.

```{r standard-example}
standard <- pepvet_preset("standard")

do.call(
  evaluate_digest,
  c(list(sequence = bsa_path, enzyme = "trypsin", missed_cleavages = 1L), standard)
)$scores
```

## DIA workflow

DIA benefits from broad peptide support across the protein because identification is not driven by one selected precursor at a time. The `dia` preset shifts weight toward coverage while preserving the same basic peptide-quality checks. [REF]

```{r dia-example}
syn_proteome <- digest_protein(syn_path, enzyme = "trypsin")
dia <- pepvet_preset("dia")

do.call(
  evaluate_digest,
  c(list(sequence = syn_path, enzyme = "trypsin", proteome = syn_proteome), dia)
)$scores
```

## Targeted workflow

Targeted assays care about a smaller set of peptides, but those peptides need to be stable and specific. The `targeted` preset pushes more weight onto `S_unique` and expects a relevant comparison proteome.

```{r targeted-example}
proteome <- digest_protein(syn_path, enzyme = "trypsin")
targeted <- pepvet_preset("targeted")

do.call(
  evaluate_digest,
  c(list(sequence = syn_path, enzyme = "trypsin", proteome = proteome), targeted)
)$scores
```

Without a proteome digest, a non-zero uniqueness weight is not valid. pepVet raises an error instead of pretending the score can be computed.

## Membrane workflow

Membrane proteins are a common reason to reject default hydrophobicity assumptions. The `membrane` preset widens the GRAVY window so that hydrophobic peptides are treated as expected rather than automatically penalized.

```{r membrane-example}
pepvet_preset("membrane")
```

This does not guarantee a membrane digest is good. It only stops the scoring model from treating all hydrophobic peptides as implausible by default.

## Degraded samples

The `ffpe_degraded` preset broadens the peptide-length window because degraded material can yield shorter but still informative peptides. That can be useful for FFPE recovery workflows and heavily processed archival material. [REF]

```{r ffpe-example}
pepvet_preset("ffpe_degraded")
```

Use this preset only when the sample context justifies it. A wider window changes the meaning of `S_length`, `S_coverage`, and `S_count`.

## Fractionated workflows

The `fractionated` preset keeps a conventional scoring profile but is useful when planning offline fractionation. The digest still needs strong peptides, but you may tolerate a broader peptide inventory because the acquisition strategy is designed to spread complexity over several runs.

## Editing a preset

Presets are plain R lists. The easiest way to customize them is to start with one and change the fields you need.

```{r modify-preset}
custom_membrane <- pepvet_preset("membrane")
custom_membrane$weights["S_coverage"] <- 0.30
custom_membrane$weights["S_hydro"] <- 0.10
if ("S_unique" %in% names(custom_membrane$weights)) {
  custom_membrane$weights["S_unique"] <- 0
}
custom_membrane$weights <- custom_membrane$weights / sum(custom_membrane$weights)

do.call(
  evaluate_digest,
  c(list(sequence = bsa_path, enzyme = "chymotrypsin-high"), custom_membrane)
)$scores
```

This is deliberate API design. Presets are data, not hidden state. You can inspect them, store them, version them, and adapt them to local methods without rewriting package internals.

## Real workflow examples

The presets are most useful in three situations.

- You need a defensible starting point for a new assay.
- You want comparisons to stay consistent across a panel of proteins.
- You need to explain why a score changed when the workflow changed.

That third case matters in practice. If a membrane target improves only after widening the GRAVY window, the result is interpretable. The score changed because the workflow assumptions changed, not because the package silently drifted.

## Summary

Use presets as starting points, not as immutable truth. They are there to make workflow assumptions explicit, reproducible, and easy to modify.

```{r session-info}
sessionInfo()
```